# Exploratory Data Analysis

## Introduction

Exploratory Data Analysis (EDA) is a critical first step in any data analysis project. In this chapter, you'll learn how to systematically explore your data to understand its structure, identify patterns, detect anomalies, and generate hypotheses for further investigation.

## The Purpose of Exploratory Data Analysis

Exploratory Data Analysis serves several important purposes in natural sciences research:

1. **Understanding Data Structure**: Gain insights into the basic properties of your dataset
2. **Checking Data Quality**: Identify missing values, outliers, and potential errors
3. **Discovering Patterns**: Detect relationships, trends, and distributions
4. **Generating Hypotheses**: Develop questions and hypotheses for formal testing
5. **Informing Analysis Choices**: Guide decisions about appropriate statistical methods

::: {.callout-tip}
## PROFESSIONAL TIP: Creating Reproducible EDA Workflows

To ensure your exploratory data analysis is reproducible and transparent:

- **Document all data transformations**: Record every cleaning step, filter, and transformation applied to raw data
- **Use R Markdown or Quarto**: Create executable documents that combine code, output, and narrative explanation
- **Version control your analysis**: Track changes to your EDA scripts using Git or similar tools
- **Save exploratory outputs**: Store key visualizations and summary statistics in organized directories
- **Create clear data lineage**: Document the origin of each dataset and how it connects to derived datasets
- **Use consistent naming conventions**: Apply systematic naming to files, variables, and functions
- **Separate exploration from confirmation**: Clearly distinguish exploratory analyses from confirmatory hypothesis testing
- **Include data validation checks**: Incorporate automated checks for data integrity and quality
- **Provide detailed method documentation**: Document statistical approaches like ANOVA types (e.g., Type II tests for unbalanced designs)
- **Share your EDA code**: Make your exploratory scripts available alongside final analyses for complete transparency
:::

## Summarizing Data

### Descriptive Statistics

Descriptive statistics provide a concise summary of your data's central tendency, dispersion, and shape:

```{r}
# Load necessary libraries
library(tidyverse)

# Load the crop yield dataset
crop_yields <- read_csv("../data/agriculture/crop_yields.csv")

# View the first few rows
head(crop_yields)

# Get summary statistics for wheat yields
wheat_summary <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`)) %>%
  summarize(
    Mean = mean(`Wheat (tonnes per hectare)`, na.rm = TRUE),
    Median = median(`Wheat (tonnes per hectare)`, na.rm = TRUE),
    StdDev = sd(`Wheat (tonnes per hectare)`, na.rm = TRUE),
    Min = min(`Wheat (tonnes per hectare)`, na.rm = TRUE),
    Max = max(`Wheat (tonnes per hectare)`, na.rm = TRUE),
    Q1 = quantile(`Wheat (tonnes per hectare)`, 0.25, na.rm = TRUE),
    Q3 = quantile(`Wheat (tonnes per hectare)`, 0.75, na.rm = TRUE)
  )

# Display the summary statistics
knitr::kable(wheat_summary, caption = "Summary Statistics for Global Wheat Yields")

# Visualize the distribution of wheat yields
ggplot(crop_yields, aes(x = `Wheat (tonnes per hectare)`)) +
  geom_histogram(bins = 30, fill = "forestgreen", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Global Wheat Yields",
       x = "Wheat Yield (tonnes per hectare)",
       y = "Frequency") +
  theme_minimal()

# Identify top wheat-producing countries (by average yield)
top_wheat_countries <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`)) %>%
  group_by(Entity) %>%
  summarize(Avg_Yield = mean(`Wheat (tonnes per hectare)`, na.rm = TRUE)) %>%
  arrange(desc(Avg_Yield)) %>%
  head(10)

# Display the top countries
knitr::kable(top_wheat_countries, caption = "Top 10 Countries by Average Wheat Yield")
```

### Frequency Tables

Frequency tables are useful for understanding the distribution of categorical variables:

```{r}
# Let's create a categorical variable based on wheat yield levels
crop_yields_with_categories <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`)) %>%
  mutate(yield_category = case_when(
    `Wheat (tonnes per hectare)` < 2 ~ "Low",
    `Wheat (tonnes per hectare)` >= 2 & `Wheat (tonnes per hectare)` < 4 ~ "Medium",
    `Wheat (tonnes per hectare)` >= 4 ~ "High"
  ))

# Frequency table for yield categories
table(crop_yields_with_categories$yield_category)

# Proportions
prop.table(table(crop_yields_with_categories$yield_category))

# Create a decade variable for temporal analysis
crop_yields_with_categories <- crop_yields_with_categories %>%
  mutate(decade = floor(Year / 10) * 10)

# Two-way frequency table: yield category by decade
yield_decade_table <- table(crop_yields_with_categories$yield_category, 
                            crop_yields_with_categories$decade)
yield_decade_table

# Convert to proportions (by row)
prop.table(yield_decade_table, margin = 1)
```

## Visualizing Distributions

### Histograms and Density Plots

Histograms and density plots help visualize the distribution of continuous variables:

```{r}
# Histogram of wheat yields
ggplot(crop_yields, aes(x = `Wheat (tonnes per hectare)`)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white", na.rm = TRUE) +
  labs(title = "Histogram of Wheat Yields", 
       x = "Wheat Yield (tonnes per hectare)", 
       y = "Frequency") +
  theme_minimal()

# Density plot
ggplot(crop_yields, aes(x = `Wheat (tonnes per hectare)`)) +
  geom_density(fill = "darkgreen", alpha = 0.5, na.rm = TRUE) +
  labs(title = "Density Plot of Wheat Yields", 
       x = "Wheat Yield (tonnes per hectare)", 
       y = "Density") +
  theme_minimal()

# Histogram with density overlay
ggplot(crop_yields, aes(x = `Wheat (tonnes per hectare)`)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "darkgreen", color = "white", na.rm = TRUE) +
  geom_density(color = "darkgreen", linewidth = 1, na.rm = TRUE) +
  labs(title = "Distribution of Wheat Yields", 
       x = "Wheat Yield (tonnes per hectare)", 
       y = "Density") +
  theme_minimal()
```

### Box Plots

Box plots are excellent for comparing distributions across groups:

```{r}
# Select a few major countries for comparison
major_wheat_producers <- c("United States", "China", "India", "Russia", "France", "Australia")

# Filter data for these countries and recent years
recent_wheat_data <- crop_yields %>%
  filter(Entity %in% major_wheat_producers, 
         Year >= 2000,
         !is.na(`Wheat (tonnes per hectare)`))

# Box plot of wheat yields by country
ggplot(recent_wheat_data, aes(x = Entity, y = `Wheat (tonnes per hectare)`)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.7) +
  labs(title = "Wheat Yields by Country (2000-present)", 
       x = "Country", 
       y = "Wheat Yield (tonnes per hectare)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Enhanced box plot with jittered points
ggplot(recent_wheat_data, aes(x = Entity, y = `Wheat (tonnes per hectare)`)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.5) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "darkgreen") +
  labs(title = "Wheat Yields by Country (2000-present)", 
       x = "Country", 
       y = "Wheat Yield (tonnes per hectare)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Bar Charts

Bar charts are useful for visualizing categorical data:

```{r}
# Calculate average wheat yield by country for the last decade
recent_avg_yields <- crop_yields %>%
  filter(Year >= 2010, !is.na(`Wheat (tonnes per hectare)`)) %>%
  group_by(Entity) %>%
  summarize(avg_wheat_yield = mean(`Wheat (tonnes per hectare)`, na.rm = TRUE)) %>%
  arrange(desc(avg_wheat_yield)) %>%
  head(10)  # Top 10 countries

# Bar chart of average wheat yields
ggplot(recent_avg_yields, aes(x = reorder(Entity, avg_wheat_yield), y = avg_wheat_yield)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(title = "Top 10 Countries by Average Wheat Yield (2010-present)", 
       x = "Country", 
       y = "Average Wheat Yield (tonnes per hectare)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Exploring Relationships

### Scatter Plots

Scatter plots help visualize relationships between two continuous variables:

```{r}
# Let's compare wheat and rice yields
crop_yields_filtered <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`), !is.na(`Rice (tonnes per hectare)`)) %>%
  filter(Year >= 2000)

# Basic scatter plot
ggplot(crop_yields_filtered, aes(x = `Wheat (tonnes per hectare)`, y = `Rice (tonnes per hectare)`)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  labs(title = "Relationship between Wheat and Rice Yields",
       x = "Wheat Yield (tonnes per hectare)", 
       y = "Rice Yield (tonnes per hectare)") +
  theme_minimal()

# Scatter plot with color by continent (we'll need to add continent information)
# For demonstration, let's create a simple mapping for a few countries
continent_mapping <- tibble(
  Entity = c("United States", "Canada", "Mexico", 
             "China", "India", "Japan", 
             "Germany", "France", "United Kingdom", 
             "Brazil", "Argentina", "Chile",
             "Egypt", "Nigeria", "South Africa",
             "Australia", "New Zealand"),
  Continent = c(rep("North America", 3), 
                rep("Asia", 3), 
                rep("Europe", 3), 
                rep("South America", 3),
                rep("Africa", 3),
                rep("Oceania", 2))
)

# Join with our dataset
crop_yields_with_continent <- crop_yields_filtered %>%
  inner_join(continent_mapping, by = "Entity")

# Scatter plot with color by continent
ggplot(crop_yields_with_continent, aes(x = `Wheat (tonnes per hectare)`, y = `Rice (tonnes per hectare)`, color = Continent)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Relationship between Wheat and Rice Yields by Continent",
       x = "Wheat Yield (tonnes per hectare)", 
       y = "Rice Yield (tonnes per hectare)") +
  theme_minimal()
```

### Correlation Analysis

Correlation analysis quantifies the strength and direction of relationships between variables:

```{r}
# Select numeric columns for correlation analysis
crop_numeric <- crop_yields %>%
  select(`Wheat (tonnes per hectare)`, `Rice (tonnes per hectare)`, `Maize (tonnes per hectare)`, `Soybeans (tonnes per hectare)`, `Potatoes (tonnes per hectare)`, `Beans (tonnes per hectare)`) %>%
  na.omit()

# Correlation matrix
cor_matrix <- cor(crop_numeric)
round(cor_matrix, 2)

# Visualize correlation matrix
library(corrplot)
corrplot(cor_matrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Crop Yields")
```

### Pair Plots

Pair plots provide a comprehensive view of relationships between multiple variables:

```{r}
# Basic pair plot
pairs(crop_numeric, pch = 19, col = "darkgreen")

# Enhanced pair plot with GGally
library(GGally)
ggpairs(crop_numeric) +
  theme_minimal() +
  labs(title = "Relationships Between Different Crop Yields")
```

## Identifying Outliers and Anomalies

### Box Plots for Outlier Detection

Box plots can help identify potential outliers:

```{r}
# Box plot to identify outliers in wheat yield
ggplot(crop_yields, aes(y = `Wheat (tonnes per hectare)`)) +
  geom_boxplot(fill = "darkgreen", alpha = 0.7, na.rm = TRUE) +
  labs(title = "Box Plot of Wheat Yields with Potential Outliers",
       y = "Wheat Yield (tonnes per hectare)") +
  theme_minimal()

# Identify potential outliers
wheat_outliers <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`)) %>%
  mutate(
    q1 = quantile(`Wheat (tonnes per hectare)`, 0.25),
    q3 = quantile(`Wheat (tonnes per hectare)`, 0.75),
    iqr = q3 - q1,
    lower_bound = q1 - 1.5 * iqr,
    upper_bound = q3 + 1.5 * iqr,
    is_outlier = `Wheat (tonnes per hectare)` < lower_bound | `Wheat (tonnes per hectare)` > upper_bound
  ) %>%
  filter(is_outlier) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`)

# Display the outliers
head(wheat_outliers, 10)
```

### Z-Scores for Outlier Detection

Z-scores can also help identify outliers:

```{r}
# Calculate z-scores for wheat yields
wheat_z_scores <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`)) %>%
  mutate(
    wheat_mean = mean(`Wheat (tonnes per hectare)`),
    wheat_sd = sd(`Wheat (tonnes per hectare)`),
    z_score = (`Wheat (tonnes per hectare)` - wheat_mean) / wheat_sd,
    is_extreme = abs(z_score) > 3
  )

# Display extreme values (z-score > 3 or < -3)
wheat_extremes <- wheat_z_scores %>%
  filter(is_extreme) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`, z_score) %>%
  arrange(desc(abs(z_score)))

head(wheat_extremes, 10)
```

## Time Series Exploration

Agricultural data often contains important temporal patterns:

```{r}
# Select a few countries for time series analysis
countries_for_ts <- c("United States", "China", "India", "France")

# Filter data for these countries
wheat_ts_data <- crop_yields %>%
  filter(Entity %in% countries_for_ts, !is.na(`Wheat (tonnes per hectare)`)) %>%
  filter(Year >= 1960)

# Time series plot
ggplot(wheat_ts_data, aes(x = Year, y = `Wheat (tonnes per hectare)`, color = Entity)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Wheat Yield Trends Over Time",
       x = "Year",
       y = "Wheat Yield (tonnes per hectare)") +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")
```

## Missing Data Analysis

Understanding patterns of missing data is crucial:

```{r}
# Check for missing values in each column
colSums(is.na(crop_yields))

# Visualize missing data patterns
if(requireNamespace("naniar", quietly = TRUE)) {
  library(naniar)
  
  # Create a visualization of missing data
  gg_miss_var(crop_yields)
  
  # Create a matrix showing missing data patterns
  vis_miss(crop_yields[, c("Entity", "Year", "Wheat (tonnes per hectare)", "Rice (tonnes per hectare)", "Maize (tonnes per hectare)")])
} else {
  message("The 'naniar' package is not installed. Install it with install.packages('naniar') to visualize missing data patterns.")
  
  # Alternative: simple summary of missing data
  missing_summary <- sapply(crop_yields, function(x) sum(is.na(x)))
  missing_df <- data.frame(
    Variable = names(missing_summary),
    Missing_Count = missing_summary,
    Missing_Percent = round(missing_summary / nrow(crop_yields) * 100, 2)
  )
  
  # Display the summary
  missing_df <- missing_df[order(-missing_df$Missing_Count), ]
  head(missing_df, 10)
}
```

## Summary

This chapter has demonstrated various techniques for exploratory data analysis using a real agricultural dataset. We've covered:

1. Computing and interpreting descriptive statistics
2. Creating and analyzing frequency tables
3. Visualizing distributions with histograms, density plots, and box plots
4. Exploring relationships with scatter plots and correlation analysis
5. Identifying outliers and anomalies
6. Analyzing time series patterns
7. Examining missing data

These techniques provide a foundation for understanding your data before proceeding to more advanced analyses. By thoroughly exploring your data, you can make informed decisions about appropriate statistical methods and generate meaningful hypotheses for testing.

## Exercises

1. Load the plant biodiversity dataset from `docs/data/ecology/biodiversity.csv` and perform a comprehensive exploratory analysis.
2. Create a histogram and density plot for another crop in the dataset. How does its distribution compare to wheat?
3. Investigate the relationship between potato yields and latitude (you'll need to find or create a dataset with latitude information).
4. Identify countries with the most significant improvement in crop yields over time.
5. Create a time series plot showing the ratio of wheat to rice yields over time for major producing countries.
6. Perform the same exploratory analyses in R for the spatial dataset in `docs/data/geography/spatial.csv`.
