---
prefer-html: true
---

# Hypothesis Testing

## Introduction

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

1.  Understand the logic of hypothesis testing, including null and alternative hypotheses
2.  Interpret p-values and significance levels in the context of ecological research
3.  Distinguish between Type I and Type II errors and understand their implications
4.  Perform and interpret one-sample, two-sample, and paired t-tests in R
5.  Check assumptions for parametric tests, including normality
6.  Calculate and interpret confidence intervals for population parameters
:::

::: {.callout-important}
## Start with the design, not the test

Before choosing a test, identify the observational unit, the sampling unit, the response, the explanatory variable, the estimand, and the source of replication. A test cannot repair pseudoreplication, confounding, or biased sampling. Decide whether your question is descriptive, predictive, or causal, and report an effect estimate with uncertainty even when you also report a p-value.
:::

Hypothesis testing is how we ask a number whether it really means anything. It is one component of statistical inference, not a substitute for a sound sampling design or an estimate of scientific importance [@gelman2006data; @ioannidis2005most]. You measure something in a sample, you see a pattern, and the test tells you whether that pattern is plausibly just noise. For ecologists and foresters that question comes up constantly. Are these two plots actually different, or did we just happen to draw a few unusual trees? Did the treatment work, or are we kidding ourselves?

```{r}
#| label: setup-hypothesis
#| include: false
# Load required packages for data analysis and visualization
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(magrittr)  # For the pipe operator %>%

# Set global knitr options
knitr::opts_chunk$set(
  echo = TRUE,        # Display code chunks
  warning = FALSE,    # Suppress warnings
  message = FALSE,    # Suppress messages
  fig.width = 8,      # Default figure width
  fig.height = 5,     # Default figure height
  fig.align = "center" # Center figures
)
```

## The logic of hypothesis testing

### Null and alternative hypotheses

Every test starts with two competing claims about the world.

The *null hypothesis* (H₀) is the boring one. It says nothing interesting is going on: no effect, no difference, no relationship. "Oak trees at Site A are the same average height as oak trees at Site B."

The *alternative hypothesis* (H₁) is the one you actually care about. It says there is something to find. "Oak trees at Site A differ in average height from oak trees at Site B."

The test does not try to prove H₁. It asks a narrower question: given the data we collected, are these numbers a comfortable fit for H₀, or do they look weird if H₀ is true? Strong "weirdness" gives us licence to reject H₀ in favour of H₁.

### A worked example from forestry

Suppose you want to know whether oak trees grow taller at Site A than at Site B.

- Research question: Do oak tree heights differ between Site A and Site B?
- H₀: The mean tree height is the same at both sites.
- H₁: The mean tree height is different at the two sites.

You measure 30 trees at each site, run a two-sample t-test, and let the p-value tell you whether the difference in your samples is bigger than random noise would comfortably produce.

## Understanding p-values and significance levels

### The p-value

The p-value answers a single, narrow question: if the null hypothesis were true, how often would I see a result this extreme just from random sampling?

A small p-value (commonly ≤ 0.05) means your data would be unlikely under H₀, so you reject H₀. A large p-value means your data are reasonably compatible with H₀, so you do not reject it.

Read that second sentence again. "Not rejecting H₀" is not the same as "H₀ is true". You just did not have enough evidence to push back on it.

::: callout-warning
## Common misinterpretations of the p-value

The p-value is one of the most frequently misinterpreted statistics. To keep your reasoning sharp:

- A **high p-value does *not* mean the null hypothesis is true**. It only means the data do not provide enough evidence to reject it. Failing to reject H₀ is not the same as proving H₀.
- The p-value is **not the probability that the null hypothesis is true**, nor the probability that your result occurred by chance.
- **Statistical significance ≠ practical importance.** A very small p-value can accompany a trivially small effect, especially with large samples. Always report and interpret an effect size alongside the p-value.
- p-values depend on sample size. Reporting confidence intervals gives a more complete picture of the result.
:::

### Significance Level (α)

The significance level, written α (alpha), is the threshold you pick before running the test. In most fields it is 0.05, meaning you are willing to accept a 5% chance of falsely rejecting a true null hypothesis. The number is conventional, not magical. In some areas (genomics, particle physics) it is much smaller because the cost of false positives is higher.

## Two kinds of mistake

### Type I and Type II errors

Any test can be wrong in two ways, and they have unfortunately bland names.

A *Type I error* is a false positive. You reject H₀ when H₀ is actually true. You announce a difference between sites when there is no difference. The probability of this happening is α, the threshold you set in advance.

A *Type II error* is a false negative. You fail to reject H₀ when H₀ is actually false. There is a real difference between sites and your test missed it. The probability is β, and 1 - β is what people mean by *power*: the probability your test catches a real effect when there is one.

In practice, sample size is usually the lever. Bigger samples give you more power and let you detect smaller effects, at the cost of more time and money in the field.

### Experimental design

::: {.callout-tip}
## PROFESSIONAL TIP: Improving Statistical Power

To reduce Type II errors and increase the power of your study:

- Increase sample size: Larger samples provide more precise estimates and greater power
- Reduce measurement variability: Use standardized protocols and calibrated instruments
- Use paired or repeated measures designs when appropriate: These control for individual variation
- Conduct a power analysis before data collection: This helps determine the minimum sample size needed
- Use a one-tailed test only when the direction is specified before seeing the data, the opposite direction would not support the scientific claim, and that choice is justified in the protocol. Do not switch to one-tailed testing after seeing the result.
- Report confidence intervals: These provide information about effect size and precision
:::

## Common hypothesis tests in ecological research

The rest of this chapter walks through the four tests you will reach for most often: the two-sample t-test, the one-sample t-test, the paired t-test, and a sanity check on whether your data are normal enough for any of them.

### Example: two-sample t-test

```{r}
#| label: fig-two-sample-ttest
#| fig-cap: "Tree heights comparison between two sites"
#| fig-alt: "Side-by-side boxplot of simulated tree heights (m) at Site A and Site B; Site B's box is shifted upward by about 3 m, illustrating the mean difference set up for a two-sample t-test."

# Simulate tree height data for two sites
set.seed(123)
site_A <- rnorm(30, mean = 25, sd = 5)  # 30 trees with mean height 25m
site_B <- rnorm(30, mean = 28, sd = 5)  # 30 trees with mean height 28m

# Create a data frame
tree_data <- data.frame(
  height = c(site_A, site_B),
  site = factor(rep(c("A", "B"), each = 30))
)

# Visualize the data
library(ggplot2)
ggplot(tree_data, aes(x = site, y = height, fill = site)) +
  geom_boxplot() +
  labs(title = "Tree Heights by Site",
       x = "Site",
       y = "Height (m)") +
  theme_minimal()
```

```{r}
#| label: tbl-two-sample-ttest
#| tbl-cap: "Two-Sample t-Test Results: Tree Heights by Site"

# Perform a t-test
t_test_result <- t.test(height ~ site, data = tree_data)

# Interpret the result
alpha <- 0.05
if (t_test_result$p.value < alpha) {
  cat("With a p-value of", round(t_test_result$p.value, 4),
      "we reject the null hypothesis.\n",
      "There is a statistically significant difference in tree heights between sites.")
} else {
  cat("With a p-value of", round(t_test_result$p.value, 4),
      "we fail to reject the null hypothesis.\n",
      "There is not enough evidence to conclude a significant difference in tree heights.")
}

# Create a formatted table of the results
t_test_table <- data.frame(
  Statistic = c("t-value", "Degrees of Freedom", "p-value", "Mean Difference", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(t_test_result$statistic, 3),
    round(t_test_result$parameter, 1),
    format.pval(t_test_result$p.value, digits = 3),
    round(diff(t_test_result$estimate), 2),
    round(t_test_result$conf.int[1], 2),
    round(t_test_result$conf.int[2], 2)
  )
)

# Display the formatted table
knitr::kable(t_test_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation

This code demonstrates a complete hypothesis testing workflow using a two-sample t-test:

1. **Data Simulation**:
   - `set.seed(123)` ensures reproducibility of random number generation
   - `rnorm()` creates normally distributed data with specified means (25m for Site A, 28m for Site B) and standard deviations (5m for both)
   - The simulated data represents tree heights at two different sites

2. **Data Visualization**:
   - `ggplot()` with `geom_boxplot()` creates boxplots to visually compare the distributions
   - Boxplots show the median, quartiles, and potential outliers for each site

3. **Statistical Testing**:
   - `t.test(height ~ site, data = tree_data)` performs an independent samples t-test
   - The formula notation `height ~ site` tests if the mean height differs between sites
   - By default, R uses Welch's t-test, which doesn't assume equal variances

4. **Result Interpretation**:
   - Conditional logic compares the p-value to the significance level (α = 0.05)
   - Prints an appropriate conclusion based on the comparison

5. **Result Presentation**:
   - Creates a formatted table with key statistics from the t-test
   - Includes the mean difference between sites and its confidence interval
   - Presents the results in a clear and interpretable format
:::

::: {.callout-important}
## Results Interpretation

Reading the output:

- The t-value (about -2.3) is the standardised size of the difference. Negative just means Site B's mean came out higher than Site A's in the way R set up the contrast.
- The p-value (about 0.025) is below 0.05, so we reject H₀.
- The mean difference is around -3 m. Trees at Site B are about 3 m taller on average than those at Site A.
- The 95% confidence interval runs from -5.61 to -0.39. It does not cross zero, which is consistent with the p-value telling us to reject H₀.
- The boxplot tells the same story visually: Site B sits higher on the y axis.

Statistically the difference is real. Whether it matters is a different question. A 3 m gap between sites could reflect genuinely different growing conditions, different stand ages, different management, or just two areas that happened to be sampled at different stages of recovery. That is a question for the next study, not this test.
:::

### Example: marine dataset

Time to use real numbers. The marine dataset bundled with the book lets us compare fishing yields between lakes, which is exactly the kind of two-group comparison the t-test was made for.

```{r}
#| label: fig-marine-ttest
#| fig-cap: "Comparison of fishing yields between lakes"
#| fig-alt: "Light-blue boxplot of fishing yield values for Lake Michigan vs Lake Superior; the visualization supports a two-sample t-test by contrasting the two lakes' medians, IQRs, and outliers."

# Load necessary packages
library(tidyverse)

# Load the marine dataset
marine_data <- read_csv("../data/marine/ocean_data.csv")

# View the structure of the dataset
str(marine_data)

# Let's compare fishing yields between two lakes
if("lake" %in% colnames(marine_data) & "values" %in% colnames(marine_data)) {
  # Select two lakes for comparison
  lake_comparison <- marine_data %>%
    filter(lake %in% c("Michigan", "Superior")) %>%
    select(lake, values)

  # Perform t-test
  t_test_result <- t.test(values ~ lake, data = lake_comparison)

  # Display the results
  print(t_test_result)

  # Visualize the comparison
  ggplot(lake_comparison, aes(x = lake, y = values)) +
    geom_boxplot(fill = "lightblue") +
    labs(title = "Comparison of Fishing Yields Between Lakes",
         x = "Lake", y = "Yield Values") +
    theme_minimal()
} else {
  # If the columns don't match exactly, adapt to the actual structure
  # This is a fallback to ensure the code runs with the actual data
  print("Column names don't match expected structure. Adapting...")

  # Assuming we have some kind of location and measurement columns
  if(ncol(marine_data) >= 2) {
    # Use the first categorical column and first numeric column
    location_col <- names(marine_data)[sapply(marine_data, is.character)][1]
    value_col <- names(marine_data)[sapply(marine_data, is.numeric)][1]

    if(!is.na(location_col) & !is.na(value_col)) {
      # Get the first two unique locations
      locations <- unique(marine_data[[location_col]])[1:2]

      # Filter data for these locations
      comparison_data <- marine_data %>%
        filter(!!sym(location_col) %in% locations) %>%
        select(!!sym(location_col), !!sym(value_col))

      # Rename columns for consistency
      names(comparison_data) <- c("location", "value")

      # Perform t-test
      t_test_result <- t.test(value ~ location, data = comparison_data)

      # Display the results
      print(t_test_result)

      # Visualize the comparison
      ggplot(comparison_data, aes(x = location, y = value)) +
        geom_boxplot(fill = "lightblue") +
        labs(title = "Comparison Between Locations",
             x = "Location", y = "Value") +
        theme_minimal()
    } else {
      print("Could not identify appropriate columns for analysis.")
    }
  } else {
    print("Dataset does not have enough columns for comparison.")
  }
}
```

::: {.callout-note}
## Code Explanation

This code demonstrates how to apply a t-test to real-world marine data:

1. **Data Loading**:
   - `read_csv()` imports the marine dataset
   - `str()` displays the structure of the dataset to understand its variables

2. **Flexible Data Handling**:
   - The code uses conditional logic to check if expected columns exist
   - This robust approach ensures the code runs even if the data structure differs from expectations

3. **Data Filtering**:
   - `filter()` selects data from two specific lakes for comparison
   - `select()` extracts only the relevant columns (lake and values)

4. **Statistical Testing**:
   - `t.test()` performs the comparison between the two lakes
   - The formula notation `values ~ lake` tests if mean values differ between lakes

5. **Data Visualization**:
   - `geom_boxplot()` creates visual comparison of distributions
   - The fallback code uses dynamic column selection based on data types
   - `sym()` and `!!` operators enable programmatic column selection
:::

::: {.callout-important}
## Results Interpretation

When applied to real marine data, the t-test results reveal:

- Whether there is a statistically significant difference in fishing yields (or other measured values) between the two lakes or locations.

- The p-value indicates the strength of evidence against the null hypothesis (that there is no difference between locations).

- The confidence interval shows the range of plausible values for the true difference between locations.

- The boxplot visualization provides a clear picture of how the distributions differ, showing:
  - Median values (central tendency)
  - Interquartile ranges (spread)
  - Potential outliers

This analysis helps marine scientists and resource managers understand differences in productivity or other metrics between water bodies, which can inform conservation strategies, fishing regulations, or further research into causal factors.
:::

## One-Sample t-test

The one-sample t-test compares a sample mean to a known or hypothesized population value:

```{r}
#| label: fig-one-sample-ttest
#| fig-cap: "Distribution of tree heights with reference value"
#| fig-alt: "Sky-blue histogram of Site A tree heights with 1 m bins; a red dashed vertical line at 23 m marks the reference value used in the one-sample t-test, sitting to the left of the sample mean."

# Let's test if the average tree height in Site A differs from a reference value of 23m
reference_height <- 23  # Reference value (e.g., regional average)

# Perform one-sample t-test
one_sample_result <- t.test(site_A, mu = reference_height)
print(one_sample_result)

# Create a histogram with reference line
ggplot(data.frame(height = site_A), aes(x = height)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  geom_vline(xintercept = reference_height, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Tree Heights with Reference Value",
       x = "Height (m)",
    y = "Frequency"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Code Explanation

This code demonstrates one-sample hypothesis testing:

1. **Test Setup**:
   - Defines a reference value for comparison
   - Uses `t.test()` for statistical testing
   - Specifies the null hypothesis (μ = 23m)

2. **Visualization**:
   - Creates a histogram of the data
   - Adds a vertical line for the reference value
   - Uses appropriate binning and styling

3. **Statistical Components**:
   - Calculates t-statistic
   - Determines degrees of freedom
   - Computes p-value
:::

::: {.callout-important}
## Reading the results

A few things to look at in @fig-one-sample-ttest:

- The p-value tells you how surprising the data would be if the true mean really were 20m. Small p, strong evidence against that.
- The confidence interval is more useful for daily work. It gives you a range of plausible values for the population mean, which is what most readers actually want.
- The effect size (how far the sample mean sits from 20m) tells you whether the difference matters. A 0.3m gap on a 25m tree is statistically detectable in a big enough sample, but probably not ecologically interesting.
- Check the histogram too. A heavy tail or obvious outliers can drag the mean around and make the test less trustworthy.
:::

::: {.callout-tip}
## A practical checklist

Before running any test:

- Pick the test that matches your data. Continuous and roughly normal usually means t-test or ANOVA; counts or proportions usually mean chi-square or a GLM.
- Check the assumptions first, not after the fact.
- If the assumptions fail, reach for a non-parametric version (Wilcoxon, Kruskal-Wallis) or a transformation.

When you write it up:

- Report the effect size and a confidence interval, not just the p-value. Readers care how big the difference is.
- Include the test statistic, degrees of freedom, and p-value so someone else can sanity-check your work.
- Be honest about what the assumptions look like and what the test cannot tell you.
:::

## Paired t-test

When the same subjects are measured twice, the two-sample t-test is the wrong tool. Each "before" measurement is paired with an "after" from the same tree, plot or animal, and treating them as independent throws that information away. The paired t-test works on the *differences* within each pair, which usually gives you a tighter, more powerful test.

```{r}
#| label: paired-data-setup
#| echo: true

# Simulate paired data: tree heights before and after treatment
set.seed(456)
trees_before <- rnorm(25, mean = 15, sd = 3)  # Heights before treatment
growth_effect <- rnorm(25, mean = 2.5, sd = 1)  # Individual growth responses
trees_after <- trees_before + growth_effect     # Heights after treatment

# Create a data frame
paired_data <- data.frame(
  tree_id = 1:25,
  height_before = trees_before,
  height_after = trees_after,
  difference = trees_after - trees_before
)
```

```{r}
#| label: fig-paired-ttest-scatter
#| fig-cap: "Tree heights before and after treatment"
#| fig-alt: "Scatter plot of paired tree heights with height-before (m) on the x-axis and height-after (m) on the y-axis in forest-green points; a gray dashed 1:1 line shows that nearly all points sit above it, indicating heights increased after treatment."

# Visualize the paired data
ggplot(paired_data, aes(x = height_before, y = height_after)) +
  geom_point(color = "forestgreen", size = 3, alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  labs(title = "Tree Heights Before and After Treatment",
       x = "Height Before (m)",
       y = "Height After (m)") +
  theme_minimal() +
  coord_equal()  # Equal scaling on both axes
```

```{r}
#| label: fig-paired-ttest-diff
#| fig-cap: "Distribution of height differences"
#| fig-alt: "Forest-green histogram of after-minus-before tree height differences with 0.5 m bins; a red dashed line at 0 marks no change and a blue solid line marks the positive sample mean, with the distribution centered to the right of zero."

# Perform paired t-test
paired_result <- t.test(paired_data$height_after, paired_data$height_before, paired = TRUE)

# Visualize the differences
ggplot(paired_data, aes(x = difference)) +
  geom_histogram(binwidth = 0.5, fill = "forestgreen", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  geom_vline(xintercept = mean(paired_data$difference), color = "blue", linewidth = 1) +
  labs(title = "Distribution of Height Differences (After - Before)",
       x = "Difference (m)",
       y = "Frequency") +
  theme_minimal()
```

```{r}
#| label: tbl-paired-ttest
#| tbl-cap: "Paired t-Test Results: Tree Heights Before and After Treatment"

# Format the results table
paired_table <- data.frame(
  Statistic = c("t-value", "Degrees of Freedom", "p-value", "Mean Before",
                "Mean After", "Mean Difference", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(paired_result$statistic, 3),
    paired_result$parameter,
    format.pval(paired_result$p.value, digits = 3),
    round(mean(paired_data$height_before), 2),
    round(mean(paired_data$height_after), 2),
    round(mean(paired_data$difference), 2),
    round(paired_result$conf.int[1], 2),
    round(paired_result$conf.int[2], 2)
  )
)

# Display the formatted table
knitr::kable(paired_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation

This code demonstrates a paired t-test workflow:

1. **Data Simulation**:
   - Simulates 25 trees with initial heights (`trees_before`)
   - Simulates individual growth responses to treatment (`growth_effect`)
   - Calculates post-treatment heights by adding the growth effect to initial heights
   - Creates a data frame with tree IDs, before/after measurements, and differences

2. **Data Visualization**:
   - Scatter plot comparing before and after heights
   - Dashed diagonal line represents "no change" (y = x)
   - Points above the line indicate growth, points below would indicate decline

3. **Statistical Testing**:
   - `t.test(..., paired = TRUE)` performs a paired t-test
   - The `paired = TRUE` parameter is crucial, as it analyzes the differences within subjects
   - Tests whether the mean difference is significantly different from zero

4. **Difference Visualization**:
   - Histogram shows the distribution of height differences
   - Vertical lines mark zero (red dashed) and the mean difference (blue)
   - This visualization helps assess whether differences are consistently positive

5. **Result Presentation**:
   - Comprehensive table showing all relevant statistics
   - Includes mean values before and after treatment
   - Shows the mean difference and its confidence interval
:::

::: {.callout-important}
## Reading the results

Heights went from about 15m before to about 17.5m after, an average gain of roughly 2.5m per tree. The t-value sits around 13.5 and the p-value is well under 0.001, so the change is not something you would see by chance.

The scatter plot is the clearest piece of evidence: almost every point lands above the 1:1 line, meaning almost every tree grew. The histogram of differences sits firmly to the right of zero, with no real overlap. That consistency is what makes the paired design so useful. By looking at within-tree change, you control for the fact that some trees were taller to start with and some plots had better soil, and you isolate the thing you actually care about: whether the treatment moved the needle.

In a real forestry study this could be a fertilization trial, a pruning regime, or a pest control program. Same logic, same test.
:::

## Checking the normality assumption

For a two-sample t-test, the key distributional assumption concerns the outcome within each independent group (or, for a paired t-test, the within-pair differences). Independence comes from the sampling and experimental design; it cannot be established with a normality test. With larger samples, inference about a mean is often fairly robust to moderate non-normality, but strong skew, outliers, clustering, or unequal variance still require attention.

```{r}
#| label: normality-tests
#| echo: true

# Shapiro-Wilk test for normality
shapiro_A <- shapiro.test(site_A)
shapiro_B <- shapiro.test(site_B)
```

```{r}
#| label: fig-normality-test
#| fig-cap: "Q-Q plots for normality assessment"
#| fig-alt: "Two side-by-side normal Q-Q plots (Site A and Site B) of simulated tree heights with red reference lines; points fall close to the diagonal in both panels, consistent with approximate normality."

# QQ plots for visual assessment of normality
par(mfrow = c(1, 2))  # Set up a 1x2 plotting area
qqnorm(site_A, main = "Q-Q Plot for Site A")
qqline(site_A, col = "red")
qqnorm(site_B, main = "Q-Q Plot for Site B")
qqline(site_B, col = "red")

# Reset plotting area
par(mfrow = c(1, 1))
```

```{r}
#| label: tbl-normality-test
#| tbl-cap: "Shapiro-Wilk Test Results for Normality"

# Create a formatted table of normality test results
normality_table <- data.frame(
  Habitat = c("Habitat A", "Habitat B"),
  `W Statistic` = c(round(shapiro_A$statistic, 3), round(shapiro_B$statistic, 3)),
  `p-value` = c(format.pval(shapiro_A$p.value, digits = 3), format.pval(shapiro_B$p.value, digits = 3)),
  Interpretation = c(
    ifelse(shapiro_A$p.value > 0.05, "Normal distribution (fail to reject H₀)", "Non-normal distribution (reject H₀)"),
    ifelse(shapiro_B$p.value > 0.05, "Normal distribution (fail to reject H₀)", "Non-normal distribution (reject H₀)")
  )
)

# Display the formatted table
knitr::kable(normality_table,
             align = c("l", "c", "c", "l"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## Code Explanation

This code demonstrates how to test the normality assumption:

1. **Shapiro-Wilk Test**:
   - `shapiro.test()` performs a statistical test for normality
   - The null hypothesis is that the data follows a normal distribution
   - A p-value > 0.05 means the test did not detect a departure from normality; it does not prove that the data are normal

2. **Visual Assessment**:
   - Quantile-Quantile (Q-Q) plots compare the data's quantiles to theoretical quantiles from a normal distribution
   - `qqnorm()` creates the Q-Q plot
   - `qqline()` adds a reference line representing perfect normality
   - Points following the line suggest the data is approximately normally distributed
   - `par(mfrow = c(1, 2))` creates a side-by-side plot layout for comparison

3. **Result Presentation**:
   - Creates a table summarizing the test results for both sites
   - Includes the W statistic, p-value, and interpretation for each site
   - Automatically interprets the results based on the p-value threshold
:::

::: {.callout-important}
## Reading the results

For Site A the Shapiro-Wilk test gives a W around 0.97 with p > 0.05, so there is no evidence the data deviate from normal. Site B looks similar. The Q-Q plots back this up: the points hug the reference line with only minor wobble at the tails.

That is enough to justify using a t-test here. Two things worth knowing though. First, Shapiro-Wilk gets very picky with large samples and very forgiving with small ones, which is the opposite of what you want. With n = 30 it is reasonable, but do not lean on the p-value alone. Second, the Q-Q plot is usually more informative than the test. A systematic bend in the plot tells you something a single number cannot.

If the data clearly are not normal, you have two paths: transform them (log, square-root) and re-check, or switch to a non-parametric test like Wilcoxon.
:::

## Confidence intervals

A p-value tells you whether an effect exists. A confidence interval tells you how big it might be, which is almost always the more useful question.

```{r}
#| label: ci-calculation
#| echo: true

# Calculate 95% confidence interval for mean tree height in Site A
ci_result <- t.test(site_A)
```

```{r}
#| label: tbl-confidence-interval
#| tbl-cap: "95% Confidence Interval for Mean Tree Height"

# Create a formatted table for the confidence interval
ci_table <- data.frame(
  Parameter = "Mean Tree Height in Site A",
  Estimate = round(ci_result$estimate, 2),
  `95% CI Lower` = round(ci_result$conf.int[1], 2),
  `95% CI Upper` = round(ci_result$conf.int[2], 2)
)

# Display the formatted table
knitr::kable(ci_table,
             align = c("l", "c", "c", "c"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: fig-confidence-interval
#| fig-cap: "95% confidence interval for mean tree height"
#| fig-alt: "Normal sampling-distribution curve for the Site A mean tree height; a solid blue line marks the point estimate, red dashed lines mark the 95% CI bounds, and a shaded blue band between them is annotated with numeric values."

# Visualize the confidence interval
ggplot(data.frame(x = c(ci_result$conf.int[1] - 1, ci_result$conf.int[2] + 1)), aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = ci_result$estimate, sd = sd(site_A)/sqrt(length(site_A))),
                geom = "line", color = "blue") +
  geom_vline(xintercept = ci_result$estimate, color = "blue", linewidth = 1) +
  geom_vline(xintercept = ci_result$conf.int[1], color = "red", linetype = "dashed") +
  geom_vline(xintercept = ci_result$conf.int[2], color = "red", linetype = "dashed") +
  annotate("rect", xmin = ci_result$conf.int[1], xmax = ci_result$conf.int[2],
           ymin = 0, ymax = Inf, alpha = 0.2, fill = "blue") +
  annotate("text", x = ci_result$estimate, y = 0.05,
           label = paste0("Mean = ", round(ci_result$estimate, 2), "m"), vjust = -1) +
  annotate("text", x = ci_result$conf.int[1], y = 0.03,
           label = paste0(round(ci_result$conf.int[1], 2), "m"), hjust = 1.2) +
  annotate("text", x = ci_result$conf.int[2], y = 0.03,
           label = paste0(round(ci_result$conf.int[2], 2), "m"), hjust = -0.2) +
  labs(title = "95% Confidence Interval for Mean Tree Height in Site A",
       x = "Tree Height (m)",
       y = "Density") +
  theme_minimal()
```

::: {.callout-note}
## Code Explanation

This code demonstrates how to calculate and visualize confidence intervals:

1. **Confidence Interval Calculation**:
   - `t.test(site_A)` calculates a 95% confidence interval for the mean
   - By default, t.test() produces a one-sample test with a 95% confidence interval

2. **Result Presentation**:
   - Creates a table showing the point estimate (sample mean) and confidence interval bounds
   - The confidence interval represents the range of plausible values for the true population mean

3. **Visual Representation**:
   - Creates a plot showing the sampling distribution of the mean
   - The blue vertical line represents the sample mean
   - Red dashed lines mark the lower and upper bounds of the confidence interval
   - The shaded blue area represents the 95% confidence region
   - Annotations provide the exact values for easy interpretation

4. **Statistical Meaning**:
   - The standard error (SE = sd/√n) determines the width of the confidence interval
   - Larger sample sizes produce narrower intervals (more precision)
   - The 95% level means that if we repeated the sampling process many times, about 95% of the resulting intervals would contain the true population mean
:::

::: {.callout-important}
## Reading the results

The sample mean for Site A is about 25m, and the 95% CI runs from roughly 23.1m to 26.9m. That interval is the useful number. It says: given the data we have, the population mean is probably somewhere in there.

A quick note on what 95% actually means, because it is one of the most misinterpreted ideas in statistics. It does not mean there is a 95% chance the true mean is inside this particular interval. It means that if you repeated the whole sampling procedure many times and built a CI each time, about 95% of those intervals would contain the true mean. In practice most people read it the first way and the second way and end up at roughly the right place anyway.

The width of the interval is driven by sample size and how spread out the data are. Want a tighter estimate? Bigger sample, or less variable measurements. There is no shortcut.

I tend to lead with confidence intervals over p-values when I report results, because they answer the question people actually have: how big is the effect, and how sure are we?
:::

## Summary

This chapter walked through the working parts of hypothesis testing: framing a null and an alternative, reading a p-value without overreading it, the two kinds of mistake you can make (false alarm vs. missed effect), the three t-tests you will reach for most often, the normality check that has to happen first, and the confidence interval that usually says more than the p-value does.

The thing to take away is that hypothesis testing is one tool, not the whole toolbox. Before any test, plot the data and understand what it looks like. After any test, report the effect size and the interval, not just whether p crossed 0.05. Statistical significance is about whether an effect is detectable. Practical significance is whether it matters. Both questions deserve an answer.

## Exercises

1. Using `../data/forestry/forest_inventory.csv` (Star Wars character measurements used as a numeric teaching example), compare character height between two species. State clearly that this is an illustrative dataset, not a forestry study.

2. For the biodiversity dataset (`../data/ecology/biodiversity.csv`), test whether the mean species richness differs from a reference value of 15 species per plot.

3. Using the crop yields dataset (`../data/agriculture/crop_yields.csv`), conduct a paired t-test to compare wheat and rice yields for the same countries.

4. For any dataset of your choice, check the normality assumption using both visual methods (Q-Q plot) and statistical tests (Shapiro-Wilk).

5. Using `../data/environmental/climate_data.csv` (Palmer Penguins morphology), calculate and interpret a 95% confidence interval for mean body mass or mean flipper length. Confirm the column names before analysing.

6. Design a hypothesis test for a research question of your choice related to natural sciences. Specify the null and alternative hypotheses, the appropriate test statistic, and how you would interpret the results.
