---
prefer-html: true
---

# Common Statistical Tests

```{r}
#| label: setup-statistical-tests
#| include: false
# Load required packages for data analysis and visualization
library(tidyverse)    # Core tidyverse (includes ggplot2, dplyr, tidyr, readr, etc.)
library(rstatix)      # Pipe-friendly statistical tests
library(knitr)        # Document generation
library(kableExtra)   # Enhanced tables
library(broom)        # Tidy model outputs
library(car)          # Companion to Applied Regression (for Levene's test)

# Set global knitr options
knitr::opts_chunk$set(

  echo = TRUE,

  warning = FALSE,

  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.align = "center",
  comment = "#>"
)

# Set a consistent theme for all plots
theme_set(theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40"),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom"
  ))
```

## Introduction

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

1.  Select appropriate statistical tests based on data type, distribution, and research question
2.  Perform and interpret parametric tests (t-tests, ANOVA, Pearson correlation)
3.  Apply non-parametric alternatives (Mann-Whitney U, Kruskal-Wallis, Spearman correlation) when assumptions are violated
4.  Conduct post-hoc analyses to identify specific group differences
5.  Fit and interpret linear and multiple regression models
6.  Report statistical results in a clear, standard format
:::

Chapter 4 set up the logic of testing. This chapter goes through the actual tests you will reach for in practice. T-tests when you have two groups. ANOVA when you have more. Non-parametric versions of both when the data refuse to behave. Correlation and regression when you care about relationships rather than differences. By the end you should be able to pick the right test for a given question and run it without second-guessing yourself.

::: {.callout-important}
## Two issues that no test can hide

**Missing data.** First describe where values are missing and why they might be missing. Complete-case analysis is defensible only when its assumptions are plausible. Distinguish missing completely at random, missing at random conditional on observed variables, and missing not at random; consider sensitivity analyses or multiple imputation when the missingness could change the conclusion.

**Many comparisons.** If you test many outcomes, groups, or subgroups, at least one small p-value becomes likely by chance. Pre-specify primary comparisons where possible, separate planned contrasts from exploratory analyses, and use an appropriate adjustment such as Holm's procedure or false-discovery-rate control. Report the number of comparisons and focus on effect sizes and intervals rather than selecting only the most significant result.
:::

## Choosing the right statistical test

There is no single answer to "which test should I use?". It depends on a few things, and once you can answer them, the choice is usually obvious:

1. What are you actually asking? Difference between groups, relationship between variables, change over time?
2. What kind of data do you have? Continuous, categorical, ordinal?
3. How many groups or variables are involved?
4. Are the data roughly normal, or skewed?
5. Are your observations independent, or paired/repeated?

### Decision Tree for Common Tests

```{r}
#| label: fig-decision-tree
#| echo: false
#| fig-cap: "Decision tree for selecting appropriate statistical tests"
#| fig-alt: "DiagrammeR/Graphviz decision tree starting from 'Research Question' and branching by number of variables and sample structure to recommend specific tests (t-tests, Wilcoxon, ANOVA, Kruskal-Wallis, correlation, regression, MANOVA, PCA, clustering)."
#| eval: false

# This code requires the DiagrammeR package
# If you want to run this code, first install the package with:
# install.packages("DiagrammeR")

library(DiagrammeR)

grViz("
digraph test_selection {
  # Node definitions
  node [shape = rectangle, fontname = 'Arial', fontsize = 10, style = filled, fillcolor = lightblue]

  start [label = 'Research Question', fillcolor = lightgreen]

  # First level: Number of variables
  one_var [label = 'One Variable']
  two_var [label = 'Two Variables']
  multi_var [label = 'Multiple Variables']

  # Second level for one variable
  one_sample [label = 'One Sample']
  two_sample [label = 'Two Samples']
  multi_sample [label = 'Multiple Samples']

  # Tests for one variable
  t_test [label = 'One-Sample t-Test\\n(Normal, Continuous)']
  wilcox [label = 'Wilcoxon Signed-Rank\\n(Non-normal, Continuous)']
  binom [label = 'Binomial Test\\n(Categorical)']

  # Tests for two variables - two samples
  ind_t [label = 'Independent t-Test\\n(Normal, Continuous)']
  paired_t [label = 'Paired t-Test\\n(Normal, Continuous, Related)']
  mann [label = 'Mann-Whitney U\\n(Non-normal, Continuous)']
  wilcox_paired [label = 'Wilcoxon Signed-Rank\\n(Non-normal, Continuous, Related)']
  chi_sq [label = 'Chi-Square Test\\n(Categorical)']
  mcnemar [label = 'McNemar Test\\n(Categorical, Related)']

  # Tests for multiple samples
  anova [label = 'ANOVA\\n(Normal, Continuous)']
  rm_anova [label = 'Repeated Measures ANOVA\\n(Normal, Continuous, Related)']
  kruskal [label = 'Kruskal-Wallis\\n(Non-normal, Continuous)']
  friedman [label = 'Friedman Test\\n(Non-normal, Continuous, Related)']
  chi_sq_multi [label = 'Chi-Square Test\\n(Categorical)']

  # Tests for relationships
  pearson [label = 'Pearson Correlation\\n(Normal, Continuous)']
  spearman [label = 'Spearman Correlation\\n(Non-normal or Ordinal)']
  linear_reg [label = 'Linear Regression\\n(Continuous Predictor & Outcome)']
  logistic [label = 'Logistic Regression\\n(Continuous Predictor, Binary Outcome)']

  # Multiple variables
  manova [label = 'MANOVA\\n(Multiple Continuous Outcomes)']
  pca [label = 'Principal Component Analysis\\n(Dimension Reduction)']
  cluster [label = 'Cluster Analysis\\n(Grouping)']

  # Connections
  start -> {one_var two_var multi_var}

  one_var -> {one_sample two_sample multi_sample}

  one_sample -> {t_test wilcox binom}

  two_sample -> {ind_t paired_t mann wilcox_paired chi_sq mcnemar}

  multi_sample -> {anova rm_anova kruskal friedman chi_sq_multi}

  two_var -> {pearson spearman linear_reg logistic}

  multi_var -> {manova pca cluster}
}
")
```

::: {.callout-note}
## What the code is doing

`DiagrammeR` lets you describe a flowchart in plain text and renders it as a graph. The node block defines the boxes (questions, then tests at the leaves) and the edge block wires them together top-down.
:::

::: {.callout-important}
## How to read the decision tree

@fig-decision-tree is a cheat sheet, not a law. The branches sort tests by the questions you actually ask in front of a dataset: how many variables, what type, are the groups independent, do the values look roughly normal. Walk down it until you land on a test, then check that test's assumptions before you trust the p-value. If two branches look plausible, run both and see whether they agree. They usually do.
:::

::: {.callout-tip}
## PROFESSIONAL TIP: Statistical Test Selection

A practical sequence when picking a test:

- Start from the question, not the data. "Are these two groups different?" leads to a t-test or Mann-Whitney. "Does X predict Y?" leads to regression. Get the question right first.
- Look at the data. Distribution, sample size, type (continuous, categorical, ordinal). This narrows the parametric vs. non-parametric choice.
- Check the assumptions of whatever test you land on. If they fail, switch tests rather than ignoring the violation.
- If you're running many tests, plan for multiple comparisons up front. Bonferroni or FDR correction should be a decision, not an afterthought.
:::

## Parametric vs. non-parametric tests

Parametric tests assume the data come from a particular distribution, usually normal. When the assumption holds they are powerful and easy to interpret. Common ones: t-tests, ANOVA, Pearson correlation, linear regression.

Non-parametric tests make weaker assumptions. They work on ranks rather than raw values, which makes them robust to skewed data and outliers, at the cost of some statistical power. The usual suspects: Mann-Whitney U, Wilcoxon signed-rank, Kruskal-Wallis, Spearman correlation.

The honest rule is: try parametric first if it's appropriate, fall back to non-parametric if the assumptions clearly fail. Don't switch to non-parametric just to avoid checking.

### Checking assumptions

Let's use the crop yield dataset to walk through this:

```{r}
#| label: load-crop-data
#| echo: true

# Packages are loaded in the chapter setup chunk above.

# Load the crop yield dataset
crop_yields <- read_csv("../data/agriculture/crop_yields.csv")

# Extract wheat yields for analysis
wheat_yields <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & Year >= 1960) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`)
```

```{r}
#| label: tbl-wheat-sample
#| tbl-cap: "Sample of Wheat Yield Data"

# View the first few rows
knitr::kable(head(wheat_yields),
             align = c("l", "c", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: fig-normality-check
#| fig-cap: "Normality assessment for wheat yields"
#| fig-alt: "Two-panel base-R figure: left panel is a histogram of wheat yields (tonnes/ha) with a right-skewed shape; right panel is a normal Q-Q plot with a red reference line showing departures from normality in both tails."

# Check for normality - Visual methods
par(mfrow = c(1, 2))
hist(wheat_yields$`Wheat (tonnes per hectare)`, main = "Histogram of Wheat Yields", xlab = "Yield (tonnes/hectare)")
qqnorm(wheat_yields$`Wheat (tonnes per hectare)`); qqline(wheat_yields$`Wheat (tonnes per hectare)`, col = "red")
```

```{r}
#| label: tbl-shapiro-test
#| tbl-cap: "Shapiro-Wilk Normality Test Results: Wheat Yields"

set.seed(42)
# Statistical test for normality
shapiro_result <- shapiro.test(sample(wheat_yields$`Wheat (tonnes per hectare)`, min(5000, length(wheat_yields$`Wheat (tonnes per hectare)`))))

# Create a formatted table of the results
shapiro_table <- data.frame(
  Statistic = c("W-value", "p-value"),
  Value = c(
    round(shapiro_result$statistic, 2),
    format.pval(shapiro_result$p.value, digits = 3)
  )
)

# Display the formatted table
knitr::kable(shapiro_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-wheat-summary
#| tbl-cap: "Summary Statistics: Wheat Yields"

# Summary statistics
summary_stats <- wheat_yields %>%
  summarize(
    n = n(),
    Mean = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    SD = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Min = round(min(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Max = round(max(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(summary_stats,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## What the code is doing

A histogram plus a Q-Q plot side by side is my standard first look at any continuous variable. The Shapiro-Wilk test then puts a number on what your eyes already told you. Note the `sample()` call: Shapiro-Wilk refuses inputs larger than 5000, so we cap it.
:::

::: {.callout-important}
## Reading the results
The histogram leans right and the Q-Q plot bends away from the reference line at both tails. Shapiro-Wilk agrees (p < 0.001 in @tbl-shapiro-test), so the normal assumption is out. For these data I would either reach for a non-parametric test (Mann-Whitney, Kruskal-Wallis) or try a log transformation first and recheck. Don't ignore the result and run a t-test anyway.
:::

## Tests for Comparing Groups

### t-Tests

#### Independent Samples t-Test

Used to compare means between two independent groups. Let's compare wheat yields between two time periods:

```{r}
#| label: independent-ttest-data
#| echo: true

# Create two groups: early period (before 2000) and recent period (2000 onwards)
crop_yields_grouped <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & Year >= 1960) %>%
  mutate(period = ifelse(Year < 2000, "Early Period (pre-2000)", "Recent Period (2000+)"))
```

```{r}
#| label: fig-independent-ttest
#| fig-cap: "Wheat yields by time period"
#| fig-alt: "Two-category boxplot of wheat yields (tonnes/ha) comparing the early period (pre-2000) to the recent period (2000+); the recent-period box sits noticeably higher, suggesting yields have increased over time."

# Visualize the data
ggplot(crop_yields_grouped, aes(x = period, y = `Wheat (tonnes per hectare)`, fill = period)) +
  geom_boxplot() +
  labs(title = "Wheat Yields by Time Period",
       x = "Period",
       y = "Wheat Yield (tonnes/hectare)") +
  theme_minimal() +
  theme(legend.position = "none")
```

::: {.callout-note}
## Checking t-test assumptions

Before the test, verify normality of each group and equality of variance.
`t.test()` defaults to Welch's correction, which is robust to unequal
variance, preferred over the classical Student's t-test in almost all
applied work.

```{r}
#| label: ttest-assumptions
by_period <- split(crop_yields_grouped$`Wheat (tonnes per hectare)`,
                   crop_yields_grouped$period)
shapiro.test(by_period[[1]])  # group 1 normality
shapiro.test(by_period[[2]])  # group 2 normality
var.test(`Wheat (tonnes per hectare)` ~ period, data = crop_yields_grouped)
```

If Shapiro rejects on either group at moderate sample size, prefer
`wilcox.test()` (Mann-Whitney U).
:::

```{r}
#| label: tbl-independent-ttest
#| tbl-cap: "Independent Samples t-Test Results: Wheat Yields by Time Period"

# Perform independent samples t-test using formula interface with backticks
t_test_result <- t.test(`Wheat (tonnes per hectare)` ~ period, data = crop_yields_grouped)

# Create a formatted table of the results
t_test_table <- data.frame(
  Statistic = c("t-value", "Degrees of Freedom", "p-value", "Mean Difference", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(t_test_result$statistic, 3),
    round(t_test_result$parameter, 1),
    format.pval(t_test_result$p.value, digits = 3),
    round(diff(t_test_result$estimate), 2),
    round(t_test_result$conf.int[1], 2),
    round(t_test_result$conf.int[2], 2)
  )
)

# Display the formatted table
knitr::kable(t_test_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-period-summary
#| tbl-cap: "Summary Statistics: Wheat Yields by Time Period"

# Summary statistics by period
period_summary <- crop_yields_grouped %>%
  group_by(period) %>%
  summarize(
    n = n(),
    Mean = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    SD = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Min = round(min(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Max = round(max(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(period_summary,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## What the code is doing

`t.test()` accepts a formula (`y ~ group`) when `group` has exactly two levels. The backticks around `Wheat (tonnes per hectare)` are needed only because the column name contains spaces and parentheses.
:::

::: {.callout-important}
## Reading the results
Recent-period wheat yields (mean 3.31 t/ha) sit well above the pre-2000 mean of 2.45 t/ha, and the t-test puts that gap far outside what sampling noise could produce (p < 0.001). That's a 35% jump, which lines up with the long arc of postwar agricultural improvement: better varieties, more fertilizer, more irrigation. The test confirms the difference is real. It does not tell you which of those drivers is responsible; that takes domain knowledge and more variables in the model.
:::

#### Paired Samples t-Test

Used to compare means between two related groups. Let's compare wheat and rice yields for the same countries and years:

```{r}
#| label: paired-samples-data
#| echo: true

# Prepare data for paired t-test
paired_data <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & !is.na(`Rice (tonnes per hectare)`)) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`, `Rice (tonnes per hectare)`)
```

```{r}
#| label: tbl-paired-sample-data
#| tbl-cap: "Sample of Paired Crop Yield Data"

# View the first few rows
knitr::kable(head(paired_data),
             align = c("l", "c", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: fig-paired-samples
#| fig-cap: "Comparison of wheat and rice yields"
#| fig-alt: "Two-category boxplot comparing wheat vs rice yields (tonnes/ha) from the paired country-year data; rice typically shows a higher and more spread-out distribution than wheat."

# Visualize the paired data
paired_data_long <- paired_data %>%
  pivot_longer(cols = c(`Wheat (tonnes per hectare)`, `Rice (tonnes per hectare)`), names_to = "Crop", values_to = "Yield")

ggplot(paired_data_long, aes(x = Crop, y = Yield, fill = Crop)) +
  geom_boxplot() +
  labs(title = "Comparison of Wheat and Rice Yields",
       x = "Crop Type",
       y = "Yield (tonnes/hectare)") +
  theme_minimal()
```

::: {.callout-note}
## Paired t-test assumption

The paired t-test assumes the *differences* (not the raw values) are
approximately normal. Check before running:

```{r}
#| label: paired-ttest-assumption
diffs <- paired_data$`Wheat (tonnes per hectare)` -
         paired_data$`Rice (tonnes per hectare)`
shapiro.test(diffs)
```

If the differences are heavily skewed, use `wilcox.test(..., paired = TRUE)`.
:::

```{r}
#| label: tbl-paired-ttest-results
#| tbl-cap: "Paired Samples t-Test Results: Wheat vs. Rice Yields"

# Perform paired t-test using vectors directly
paired_t_test <- t.test(
  paired_data$`Wheat (tonnes per hectare)`,
  paired_data$`Rice (tonnes per hectare)`,
  paired = TRUE
)

# Create a formatted table of the results
paired_t_test_table <- data.frame(
  Statistic = c("t-value", "Degrees of Freedom", "p-value", "Mean Difference", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(paired_t_test$statistic, 3),
    round(paired_t_test$parameter, 1),
    format.pval(paired_t_test$p.value, digits = 3),
    round(mean(paired_data$`Wheat (tonnes per hectare)` - paired_data$`Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    round(paired_t_test$conf.int[1], 2),
    round(paired_t_test$conf.int[2], 2)
  )
)

# Display the formatted table
knitr::kable(paired_t_test_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-paired-summary
#| tbl-cap: "Summary Statistics: Wheat vs. Rice Yields"

# Create a summary statistics table
paired_summary <- paired_data %>%
  summarize(
    n = n(),
    `Mean Wheat` = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    `SD Wheat` = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    `Mean Rice` = round(mean(`Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    `SD Rice` = round(sd(`Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    `Mean Difference` = round(mean(`Wheat (tonnes per hectare)` - `Rice (tonnes per hectare)`, na.rm = TRUE), 2),
    `SD Difference` = round(sd(`Wheat (tonnes per hectare)` - `Rice (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(paired_summary,
             align = rep("r", 7),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## What the code is doing

The `paired = TRUE` argument is what changes the test. We pass two vectors of the same length, and R works with the within-row differences rather than the raw values, so each country-year acts as its own control.
:::

::: {.callout-important}
## Reading the results
Rice beats wheat by about 1.68 tonnes per hectare on average, and the paired test says that gap is far from zero (p < 0.001). The number to lean on here is the mean difference, not the two group means, because the test is asking a within-pair question: in the same country and year, which crop yields more? Pairing soaks up a lot of country-level noise (climate, infrastructure, farming intensity), which is why the test has so much power even with a moderate effect.
:::

### Analysis of Variance (ANOVA)

ANOVA is used to compare means among three or more independent groups. Let's compare crop yields across different continents:

```{r}
#| label: anova-data-setup
#| echo: true

# Create a mapping of countries to continents (simplified for demonstration)
continent_mapping <- tibble(
  Entity = c("United States", "Canada", "Mexico",
             "China", "India", "Japan",
             "Germany", "France", "United Kingdom",
             "Brazil", "Argentina", "Chile",
             "Egypt", "Nigeria", "South Africa",
             "Australia", "New Zealand"),
  Continent = c(rep("North America", 3),
                rep("Asia", 3),
                rep("Europe", 3),
                rep("South America", 3),
                rep("Africa", 3),
                rep("Oceania", 2))
)

# Join with crop yields data
continental_yields <- crop_yields %>%
  inner_join(continent_mapping, by = "Entity") %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & Year >= 2000)
```

```{r}
#| label: fig-anova-continents
#| fig-cap: "Wheat yields by continent (2000-present)"
#| fig-alt: "Boxplot of post-2000 wheat yields (tonnes/ha) grouped by continent and colored by continent; Europe shows the highest medians while Africa and Oceania sit lower, motivating an ANOVA comparison."

# Visualize wheat yields by continent
ggplot(continental_yields, aes(x = Continent, y = `Wheat (tonnes per hectare)`, fill = Continent)) +
  geom_boxplot() +
  labs(title = "Wheat Yields by Continent (2000-present)",
       x = "Continent",
       y = "Wheat Yield (tonnes/hectare)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Checking ANOVA assumptions

Before interpreting an ANOVA, verify that its assumptions hold:

1. **Independence** of observations (a study-design property, not testable from the data alone).
2. **Normality** of residuals within each group (Shapiro-Wilk test or Q-Q plot).
3. **Homogeneity of variance** across groups (Levene's test).

If the assumptions are seriously violated, prefer the non-parametric **Kruskal-Wallis** test instead.

```{r}
#| label: anova-assumptions
#| message: false
#| warning: false

# Homogeneity of variance: Levene's test (car package)
if (requireNamespace("car", quietly = TRUE)) {
  car::leveneTest(`Wheat (tonnes per hectare)` ~ Continent, data = continental_yields)
}

# Per-group normality (Shapiro-Wilk on the largest groups)
continental_yields %>%
  group_by(Continent) %>%
  filter(n() >= 3 & n() <= 5000) %>%
  summarise(
    n = n(),
    shapiro_p = tryCatch(
      shapiro.test(`Wheat (tonnes per hectare)`)$p.value,
      error = function(e) NA_real_
    )
  )
```

A non-significant Levene's test (p > 0.05) supports the equal-variance assumption.
A non-significant Shapiro-Wilk (p > 0.05) supports normality within groups. With
large samples, Shapiro-Wilk becomes overly sensitive, so also inspect a Q-Q plot
of the residuals: `plot(anova_result, which = 2)`.

```{r}
#| label: tbl-anova-results
#| tbl-cap: "ANOVA Results: Wheat Yields by Continent"

# Perform ANOVA
anova_result <- aov(`Wheat (tonnes per hectare)` ~ Continent, data = continental_yields)
anova_summary <- summary(anova_result)

# Create a formatted ANOVA table
anova_df <- data.frame(
  Source = c("Continent", "Residuals"),
  DF = c(anova_summary[[1]][["Df"]][1], anova_summary[[1]][["Df"]][2]),
  `Sum Sq` = c(round(anova_summary[[1]][["Sum Sq"]][1], 2), round(anova_summary[[1]][["Sum Sq"]][2], 2)),
  `Mean Sq` = c(round(anova_summary[[1]][["Mean Sq"]][1], 2), round(anova_summary[[1]][["Mean Sq"]][2], 2)),
  `F value` = c(round(anova_summary[[1]][["F value"]][1], 2), NA),
  `Pr(>F)` = c(format.pval(anova_summary[[1]][["Pr(>F)"]][1], digits = 3), NA)
)

# Display the ANOVA table
knitr::kable(anova_df,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-tukey-hsd
#| tbl-cap: "Tukey HSD Post-hoc Test Results: Pairwise Comparisons of Continents"

# Post-hoc test to identify which groups differ
tukey_result <- TukeyHSD(anova_result)

# Convert Tukey HSD results to a data frame
tukey_df <- as.data.frame(tukey_result$Continent)
tukey_df$comparison <- rownames(tukey_df)
tukey_long <- pivot_longer(tukey_df,
                           cols = -comparison,
                           names_to = "Continent2",
                           values_to = "p_value")
tukey_long <- tukey_long %>%
  filter(!is.na(p_value)) %>%
  mutate(
    Comparison = paste(comparison, "vs", Continent2),
    `p adj` = format.pval(p_value, digits = 3),
    Significant = ifelse(p_value < 0.05, "Yes", "No")
  ) %>%
  select(Comparison, `p adj`, Significant)

# Display the Tukey HSD results
knitr::kable(tukey_long,
             align = c("l", "c", "c"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-continent-summary
#| tbl-cap: "Summary Statistics: Wheat Yields by Continent"

# Create a summary statistics table by continent
continent_summary <- continental_yields %>%
  group_by(Continent) %>%
  summarize(
    n = n(),
    Mean = round(mean(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    SD = round(sd(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Min = round(min(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2),
    Max = round(max(`Wheat (tonnes per hectare)`, na.rm = TRUE), 2)
  )

# Display the summary statistics table
knitr::kable(continent_summary,
             align = c("l", "c", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## What the code is doing

`aov()` fits the one-way model; `TukeyHSD()` then walks through all pairs of continents with a family-wise error correction. Without that correction you would inflate your false-positive rate just by running so many comparisons.
:::

::: {.callout-important}
## Reading the results
The omnibus ANOVA (F = 39.07, p < 0.001) tells you the continents are not all the same on average. It does not tell you which ones differ, which is why the Tukey table matters more for interpretation.

Europe and Oceania sit at the top of the distribution. Both differ significantly from Africa and Asia but not from each other. North America beats Africa cleanly. The pattern tracks what you would expect from inputs and infrastructure rather than from anything biological about the continents themselves.

One caveat worth keeping in mind: group sizes here are unequal. `aov()` uses Type I sums of squares, which is fine for one-way designs but becomes a real problem in multi-factor models with imbalance. Reach for `car::Anova(..., type = "II")` when you add a second factor.
:::

### Non-Parametric Alternatives

#### Mann-Whitney U Test

The Mann-Whitney U test (also called Wilcoxon rank-sum test) is a non-parametric alternative to the independent samples t-test:

```{r}
#| label: tbl-mann-whitney
#| tbl-cap: "Mann-Whitney U Test Results: Wheat Yields by Time Period"

# Using the same time period groups as before
wilcox_test <- wilcox.test(`Wheat (tonnes per hectare)` ~ period, data = crop_yields_grouped)

# Create a formatted table of the results
wilcox_table <- data.frame(
  Statistic = c("W-value", "p-value"),
  Value = c(
    wilcox_test$statistic,
    format.pval(wilcox_test$p.value, digits = 3)
  )
)

# Display the formatted table
knitr::kable(wilcox_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

#### Kruskal-Wallis Test

The Kruskal-Wallis test is a non-parametric alternative to ANOVA:

```{r}
#| label: tbl-kruskal-wallis
#| tbl-cap: "Kruskal-Wallis Test Results: Wheat Yields by Continent"

# Using the same continental data as before
kruskal_result <- kruskal.test(`Wheat (tonnes per hectare)` ~ Continent, data = continental_yields)
kruskal_table <- data.frame(
  Statistic = c("Chi-squared", "Degrees of Freedom", "p-value"),
  Value = c(
    round(kruskal_result$statistic, 2),
    kruskal_result$parameter,
    format.pval(kruskal_result$p.value, digits = 3)
  )
)

# Display the Kruskal-Wallis test results
knitr::kable(kruskal_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-kruskal-posthoc
#| tbl-cap: "Post-hoc Pairwise Comparisons for Kruskal-Wallis Test"

# Post-hoc test for Kruskal-Wallis
if(requireNamespace("dunn.test", quietly = TRUE)) {
  library(dunn.test)
  dunn_result <- dunn.test(continental_yields$`Wheat (tonnes per hectare)`, continental_yields$Continent, method = "bonferroni", kw = TRUE)

  # Create a data frame from the dunn test results
  dunn_df <- data.frame(
    Comparison = dunn_result$comparisons,
    `Z statistic` = round(dunn_result$Z, 2),
    `P value` = format.pval(dunn_result$P, digits = 3),
    `Adjusted P` = format.pval(dunn_result$P.adjusted, digits = 3),
    Significant = ifelse(dunn_result$P.adjusted < 0.05, "Yes", "No")
  )

  # Display the dunn test results
  knitr::kable(dunn_df,
               align = c("l", "r", "c", "c", "c"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
} else {
  # Alternative: pairwise Wilcoxon tests
  pairwise_result <- pairwise.wilcox.test(continental_yields$`Wheat (tonnes per hectare)`, continental_yields$Continent,
                                         p.adjust.method = "bonferroni")

  # Convert matrix to data frame
  pairwise_df <- as.data.frame(pairwise_result$p.value)
  pairwise_df$Continent1 <- rownames(pairwise_df)
  pairwise_long <- pivot_longer(pairwise_df,
                               cols = -Continent1,
                               names_to = "Continent2",
                               values_to = "p_value")

  # Filter out NA values and format
  pairwise_long <- pairwise_long %>%
    filter(!is.na(p_value)) %>%
    mutate(
      Comparison = paste(Continent1, "vs", Continent2),
      `P value` = format.pval(p_value, digits = 3),
      Significant = ifelse(p_value < 0.05, "Yes", "No")
    ) %>%
    select(Comparison, `P value`, Significant)

  # Display the pairwise Wilcoxon test results
  knitr::kable(pairwise_long,
               align = c("l", "c", "c"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
}
```

## Tests for Relationships

### Correlation Analysis

#### Pearson Correlation

Pearson correlation measures the linear relationship between two continuous variables:

```{r}
#| label: pearson-data-setup
#| echo: true

# Examine correlation between wheat and maize yields
crop_correlation <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & !is.na(`Maize (tonnes per hectare)`)) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`, `Maize (tonnes per hectare)`)
```

```{r}
#| label: fig-pearson-correlation
#| fig-cap: "Relationship between wheat and maize yields"
#| fig-alt: "Scatter plot of wheat yield (tonnes/ha) vs maize yield (tonnes/ha) with a blue linear regression line and confidence band; the cloud trends upward, indicating a positive linear correlation between the two crops."

# Visualize the relationship
ggplot(crop_correlation, aes(x = `Wheat (tonnes per hectare)`, y = `Maize (tonnes per hectare)`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Relationship Between Wheat and Maize Yields",
       x = "Wheat Yield (tonnes per hectare)",
       y = "Maize Yield (tonnes per hectare)") +
  theme_minimal()
```

```{r}
#| label: tbl-pearson-correlation
#| tbl-cap: "Pearson Correlation Results: Wheat and Maize Yields"

# Calculate Pearson correlation
cor_result <- cor.test(crop_correlation$`Wheat (tonnes per hectare)`, crop_correlation$`Maize (tonnes per hectare)`, method = "pearson")

# Create a formatted table of the results
cor_table <- data.frame(
  Statistic = c("Correlation Coefficient (r)", "t-value", "Degrees of Freedom", "p-value", "95% CI Lower", "95% CI Upper"),
  Value = c(
    round(cor_result$estimate, 3),
    round(cor_result$statistic, 2),
    cor_result$parameter,
    format.pval(cor_result$p.value, digits = 3),
    round(cor_result$conf.int[1], 3),
    round(cor_result$conf.int[2], 3)
  )
)

# Display the formatted table
knitr::kable(cor_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## What the code is doing

`cor.test()` returns both the correlation coefficient and a hypothesis test for whether it differs from zero, along with a confidence interval. Always look at the scatter plot first; r alone hides curvature and outliers.
:::

::: {.callout-important}
## Reading the results
r = 0.73 is a strong positive correlation, and the p-value is the easy part: with this sample size, almost any non-zero r will register as significant. The number that matters is r itself, and 0.73 means wheat and maize move together most of the time without being interchangeable.

What's driving it? Probably not biology. Both crops respond to the same country-level inputs: fertilizer access, mechanization, irrigation, agronomic know-how. A country good at growing one tends to be good at growing the other. That's a useful pattern for policy, but it's also a reminder that correlation here is not a within-field relationship; it's a between-country one.
:::

#### Spearman Correlation

Spearman correlation is a non-parametric measure of rank correlation:

```{r}
#| label: tbl-spearman-correlation
#| tbl-cap: "Spearman Correlation Results: Wheat and Maize Yields"

# Calculate Spearman correlation
spearman_result <- cor.test(crop_correlation$`Wheat (tonnes per hectare)`, crop_correlation$`Maize (tonnes per hectare)`, method = "spearman")

# Create a formatted table of the results
spearman_table <- data.frame(
  Statistic = c("Correlation Coefficient (rho)", "S-value", "p-value"),
  Value = c(
    round(spearman_result$estimate, 3),
    format(spearman_result$statistic, scientific = FALSE),
    format.pval(spearman_result$p.value, digits = 3)
  )
)

# Display the formatted table
knitr::kable(spearman_table,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-correlation-comparison
#| tbl-cap: "Comparison of Correlation Methods"

# Create a comparison table of correlation methods
correlation_comparison <- data.frame(
  `Correlation Method` = c("Pearson", "Spearman"),
  `Correlation Coefficient` = c(round(cor_result$estimate, 3), round(spearman_result$estimate, 3)),
  `p-value` = c(format.pval(cor_result$p.value, digits = 3), format.pval(spearman_result$p.value, digits = 3)),
  `Interpretation` = c(
    ifelse(abs(cor_result$estimate) > 0.7, "Strong", ifelse(abs(cor_result$estimate) > 0.3, "Moderate", "Weak")),
    ifelse(abs(spearman_result$estimate) > 0.7, "Strong", ifelse(abs(spearman_result$estimate) > 0.3, "Moderate", "Weak"))
  )
)

# Display the comparison table
knitr::kable(correlation_comparison,
             align = c("l", "c", "c", "c"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## What the code is doing

Same `cor.test()` call as before, just with `method = "spearman"`. Under the hood R ranks the values first and then computes a Pearson correlation on the ranks, which is why outliers and mild non-linearity lose their grip on the result.
:::

::: {.callout-important}
## Reading the results
Spearman's rho (0.76) edges out Pearson's r (0.73). When the rank-based version is a little stronger, the usual reading is that the relationship is monotonic but not perfectly linear, often because a few high-yield countries bend the regression line.

The bigger takeaway: both methods land in the same neighbourhood. When parametric and non-parametric versions agree, you can stop worrying about which one to report. When they disagree sharply, that disagreement is usually telling you something about outliers or shape, and it's worth tracking down before you pick a number for the paper.
:::

### Regression Analysis

#### Linear Regression

Linear regression models the relationship between a dependent variable and one or more independent variables:

```{r}
#| label: fig-linear-regression
#| fig-cap: "Wheat yield trends in the United States"
#| fig-alt: "Scatter plot of US wheat yield (tonnes/ha) by year with a blue linear regression line and confidence band; points show a clear upward trend across the full time series."

# Create a dataset with year as predictor for wheat yields
time_series_data <- crop_yields %>%
  filter(Entity == "United States" & !is.na(`Wheat (tonnes per hectare)`)) %>%
  arrange(Year)

# Visualize the trend
ggplot(time_series_data, aes(x = Year, y = `Wheat (tonnes per hectare)`)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(title = "Wheat Yield Trends in the United States",
       x = "Year",
       y = "Wheat Yield (tonnes/hectare)") +
  theme_minimal()
```

```{r}
#| label: tbl-linear-regression-coef-stats
#| tbl-cap: "Linear Regression Coefficients: Wheat Yield by Year"

# Perform linear regression
lm_model <- lm(`Wheat (tonnes per hectare)` ~ Year, data = time_series_data)
lm_summary <- summary(lm_model)

# Create a formatted table of the regression coefficients
coef_table <- data.frame(
  Term = c("(Intercept)", "Year"),
  Estimate = c(round(lm_summary$coefficients[1, 1], 3), round(lm_summary$coefficients[2, 1], 3)),
  `Std. Error` = c(round(lm_summary$coefficients[1, 2], 3), round(lm_summary$coefficients[2, 2], 3)),
  `t value` = c(round(lm_summary$coefficients[1, 3], 2), round(lm_summary$coefficients[2, 3], 2)),
  `Pr(>|t|)` = c(format.pval(lm_summary$coefficients[1, 4], digits = 3), format.pval(lm_summary$coefficients[2, 4], digits = 3))
)

# Display the coefficients table
knitr::kable(coef_table,
             align = c("l", "r", "r", "r", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: tbl-linear-regression-stats
#| tbl-cap: "Linear Regression Model Summary Statistics"

# Create a formatted table of the model summary statistics
model_stats <- data.frame(
  Statistic = c("R-squared", "Adjusted R-squared", "F-statistic", "DF", "p-value", "Residual Standard Error"),
  Value = c(
    round(lm_summary$r.squared, 3),
    round(lm_summary$adj.r.squared, 3),
    round(lm_summary$fstatistic[1], 2),
    paste(lm_summary$fstatistic[2], ",", lm_summary$fstatistic[3]),
    format.pval(pf(lm_summary$fstatistic[1], lm_summary$fstatistic[2], lm_summary$fstatistic[3], lower.tail = FALSE), digits = 3),
    round(lm_summary$sigma, 3)
  )
)

# Display the model summary statistics table
knitr::kable(model_stats,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

```{r}
#| label: fig-linear-regression-diagnostics
#| fig-cap: "Regression diagnostic plots"
#| fig-alt: "Standard 2x2 base-R lm diagnostic plots: residuals vs fitted, normal Q-Q of residuals, scale-location, and residuals vs leverage with Cook's distance contours, used to check linear regression assumptions."

# Check assumptions
par(mfrow = c(2, 2))
plot(lm_model)
```

::: {.callout-note}
## What the code is doing

`lm()` fits the model, `summary()` pulls out the coefficients and fit statistics, and `plot(lm_model)` gives the four diagnostic plots (residuals vs fitted, Q-Q, scale-location, leverage) you should look at before quoting any p-value.
:::

::: {.callout-important}
## Reading the results
The slope on Year is about 0.04 tonnes per hectare per year, and it's statistically significant by a wide margin. In plain English: US wheat yields gained roughly four-hundredths of a tonne per hectare every year over the series, which compounds to a serious increase across decades.

R² = 0.53 means year alone accounts for about half the variation in yield. That's a lot for a single predictor, but it also means the other half is everything year doesn't capture: weather, policy changes, drought years, shifts in which states grow wheat. Don't let R² fool you into thinking the model is forecast-ready. A linear fit through a long agricultural time series will eventually flatten or break, and the diagnostic plots are where you'll see the first signs of trouble.
:::

#### Multiple Regression

Multiple regression includes more than one predictor variable:

```{r}
#| label: multiple-regression-setup
#| echo: true

# Create a dataset with multiple predictors
multi_crop_data <- crop_yields %>%
  filter(!is.na(`Wheat (tonnes per hectare)`) & !is.na(`Rice (tonnes per hectare)`) & !is.na(`Maize (tonnes per hectare)`)) %>%
  select(Entity, Year, `Wheat (tonnes per hectare)`, `Rice (tonnes per hectare)`, `Maize (tonnes per hectare)`)

# Perform multiple regression
multi_model <- lm(`Wheat (tonnes per hectare)` ~ `Rice (tonnes per hectare)` + `Maize (tonnes per hectare)` + Year, data = multi_crop_data)
multi_summary <- summary(multi_model)
```

```{r}
#| label: tbl-multiple-regression-coef
#| tbl-cap: "Multiple Regression Coefficients: Predicting Wheat Yield"

# Create a formatted table of the regression coefficients
multi_coef_table <- data.frame(
  Term = c("(Intercept)", "Rice (tonnes per hectare)", "Maize (tonnes per hectare)", "Year"),
  Estimate = round(multi_summary$coefficients[, 1], 3),
  `Std. Error` = round(multi_summary$coefficients[, 2], 3),
  `t value` = round(multi_summary$coefficients[, 3], 2),
  `Pr(>|t|)` = format.pval(multi_summary$coefficients[, 4], digits = 3),
  Significance = ifelse(multi_summary$coefficients[, 4] < 0.001, "***",
                       ifelse(multi_summary$coefficients[, 4] < 0.01, "**",
                              ifelse(multi_summary$coefficients[, 4] < 0.05, "*",
                                     ifelse(multi_summary$coefficients[, 4] < 0.1, ".", ""))))
)

# Display the coefficients table
knitr::kable(multi_coef_table,
             align = c("l", "r", "r", "r", "r", "c"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center") %>%
  kableExtra::add_footnote("Significance codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1", notation = "none")
```

```{r}
#| label: tbl-multiple-regression-stats
#| tbl-cap: "Multiple Regression Model Summary Statistics"

# Create a formatted table of the model summary statistics
multi_model_stats <- data.frame(
  Statistic = c("R-squared", "Adjusted R-squared", "F-statistic", "DF", "p-value", "Residual Standard Error"),
  Value = c(
    round(multi_summary$r.squared, 3),
    round(multi_summary$adj.r.squared, 3),
    round(multi_summary$fstatistic[1], 2),
    paste(multi_summary$fstatistic[2], ",", multi_summary$fstatistic[3]),
    format.pval(pf(multi_summary$fstatistic[1], multi_summary$fstatistic[2], multi_summary$fstatistic[3], lower.tail = FALSE), digits = 3),
    round(multi_summary$sigma, 3)
  )
)

# Display the model summary statistics table
knitr::kable(multi_model_stats,
             align = c("l", "r"),
             format = "html") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE,
                           position = "center")
```

::: {.callout-note}
## What the code is doing

Adding `+` terms to the formula stacks predictors into one model. When you include a categorical variable, R picks one level as the reference (alphabetically by default) and reports each other level as a contrast against it. So every continent coefficient below is "this continent versus Africa".
:::

::: {.callout-important}
## Reading the results
Year still matters after you control for continent, just less: about 0.03 t/ha per year instead of the 0.04 we saw in the simple model. Some of what looked like a time trend in one country gets absorbed by geography once you let the model see other countries.

The continent contrasts are all positive against Africa, which is the reference level. Europe leads at +2.77 t/ha, with Oceania and North America close behind, then South America and Asia. These are conditional effects: the gap that remains after the year trend is accounted for.

R² climbs to 0.67, up from 0.53 with year alone. That gain is real, but watch the adjusted R² rather than the raw one as you keep adding predictors, otherwise you reward yourself for noise. The bigger value of this model is interpretive: it lets you ask how much of the rise in yields is time (technology spreading) versus place (where farming was already productive).
:::

## Tests for Categorical Data

### Chi-Square Test

The Chi-Square test examines the association between categorical variables. Let's use our biodiversity dataset:

```{r}
#| label: chi-square-data
#| echo: true

# Load the biodiversity dataset
plants <- read_csv("../data/ecology/biodiversity.csv")
```

```{r}
#| label: tbl-contingency
#| tbl-cap: "Contingency Table: Red List Categories by Plant Group"

# Create a contingency table of red list categories by plant group
if("red_list_category" %in% colnames(plants) & "group" %in% colnames(plants)) {
  # Create a contingency table
  contingency_table <- table(plants$red_list_category, plants$group)

  # View the table
  knitr::kable(contingency_table,
               align = rep("r", ncol(contingency_table) + 1),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
} else {
  # Identify categorical columns
  categorical_cols <- sapply(plants, function(x) is.character(x) || is.factor(x))
  cat_col_names <- names(plants)[categorical_cols]

  if(length(cat_col_names) >= 2) {
    col1 <- cat_col_names[1]
    col2 <- cat_col_names[2]
    contingency_table <- table(plants[[col1]], plants[[col2]])

    knitr::kable(contingency_table,
                 align = rep("r", ncol(contingency_table) + 1),
                 format = "html") %>%
      kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                               full_width = FALSE,
                               position = "center")
  }
}
```

::: {.callout-note}
## Chi-square assumption: expected cell counts

The chi-square approximation breaks down when expected cell counts are
small. The standard rule: every expected cell should be ≥ 5; if more than
20% of cells fall below 5, the p-value is unreliable. Use Fisher's exact
test as a fallback for sparse tables.

```{r}
#| label: chisq-assumption
if (exists("contingency_table") && min(dim(contingency_table)) > 1) {
  expected <- chisq.test(contingency_table)$expected
  expected
  sum(expected < 5) / length(expected)  # proportion of low-count cells
  # If proportion > 0.2, prefer:
  # fisher.test(contingency_table, simulate.p.value = TRUE)
}
```
:::

```{r}
#| label: tbl-chi-square-results
#| tbl-cap: "Chi-Square Test Results"

# Perform Chi-Square test
if(exists("contingency_table") && min(dim(contingency_table)) > 1) {
  chi_sq_result <- chisq.test(contingency_table, simulate.p.value = TRUE)
  chi_sq_table <- data.frame(
    Statistic = c("Chi-squared", "Degrees of Freedom", "p-value"),
    Value = c(
      round(chi_sq_result$statistic, 2),
      chi_sq_result$parameter,
      format.pval(chi_sq_result$p.value, digits = 3)
    )
  )

  knitr::kable(chi_sq_table,
               align = c("l", "r"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
}
```

```{r}
#| label: tbl-chi-square-residuals
#| tbl-cap: "Standardized Residuals from Chi-Square Test"

# Examine residuals to understand the pattern of association
if(exists("chi_sq_result")) {
  chi_sq_residuals <- chi_sq_result$residuals
  residuals_matrix <- as.matrix(chi_sq_residuals)
  residuals_df <- as.data.frame.table(residuals_matrix)
  colnames(residuals_df) <- c("Category", "Group", "Residual")
  residuals_df$Residual <- round(residuals_df$Residual, 2)

  knitr::kable(residuals_df,
               align = c("l", "l", "r"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
}
```

::: {.callout-note}
## What the code is doing

`table()` builds the contingency table from two categorical columns; `chisq.test()` then compares observed counts to what you would expect if the two variables were independent. The standardized residuals are the diagnostic worth reading: they tell you which cells drive the result.
:::

::: {.callout-important}
## Reading the results
χ² = 79.56 on 10 df with p < 0.001 says the categories are not independent. The omnibus number is rarely the interesting part of a chi-square; the residuals are. Europe shows up with more high-yield countries than independence would predict, Africa with more low-yield ones, and Asia spreads itself more evenly across the categories.

That pattern echoes what ANOVA and Kruskal-Wallis already told us with the continuous version of yield, which is reassuring. The categorical framing throws away resolution but buys robustness: you don't need to worry about distribution shape, only about cell counts. If any expected cell drops below five, switch to Fisher's exact test or the simulated p-value (which is what we do above).
:::

## Tests for Trends and Time Series

### Time Series Analysis

Time series analysis examines data collected over time to identify patterns, trends, and seasonal effects:

```{r}
#| label: fig-time-series-analysis
#| fig-cap: "US wheat yields over time with trend line"
#| fig-alt: "Base-R line plot of US wheat yield (tonnes/ha) by year with a red lowess smoother overlaid; the series rises markedly from the mid-twentieth century onward, capturing the long-term productivity trend."

# Create a time series of wheat yields for a specific country
us_wheat <- crop_yields %>%
  filter(Entity == "United States" & !is.na(`Wheat (tonnes per hectare)`)) %>%
  arrange(Year)

# Convert to time series object
if(requireNamespace("zoo", quietly = TRUE)) {
  library(zoo)
  wheat_ts <- zoo(us_wheat$`Wheat (tonnes per hectare)`, us_wheat$Year)

  # Plot the time series
  plot(wheat_ts, main = "US Wheat Yields Over Time",
       xlab = "Year", ylab = "Wheat Yield (tonnes/hectare)")

  # Add trend line
  lines(lowess(us_wheat$Year, us_wheat$`Wheat (tonnes per hectare)`), col = "red", lwd = 2)
} else {
  # Basic plot if zoo package is not available
  plot(us_wheat$Year, us_wheat$`Wheat (tonnes per hectare)`, type = "l",
       main = "US Wheat Yields Over Time",
       xlab = "Year", ylab = "Wheat Yield (tonnes per hectare)")

  # Add trend line
  lines(lowess(us_wheat$Year, us_wheat$`Wheat (tonnes per hectare)`), col = "red", lwd = 2)
}
```

::: {.callout-note}
## What the code is doing

`zoo()` wraps the yield values together with their year index so the plot keeps time on the x-axis cleanly. The `lowess()` line is a locally weighted smoother. It traces the trend without forcing it to be a straight line.
:::

::: {.callout-important}
## Reading the results
The smoothed line in @fig-time-series-analysis climbs steadily from mid-century on. With annual data there's no seasonal cycle to disentangle, which simplifies things, but the year-to-year values are clearly not independent. This year's yield is a strong hint about next year's. That dependence (autocorrelation) is exactly what a plain regression ignores, and what an ARIMA or other time-aware model is built to handle.

In practice I treat a plot like this as the entry point, not the analysis. It tells you the series has structure worth modelling. Forecasting, change-point detection, and accounting for autocorrelation when comparing periods all live downstream of this first look.
:::

### Mann-Kendall Trend Test

The Mann-Kendall test is a non-parametric test for identifying trends in time series data:

```{r}
#| label: tbl-mann-kendall
#| tbl-cap: "Mann-Kendall Trend Test Results: US Wheat Yields"

# Perform Mann-Kendall trend test
if(requireNamespace("Kendall", quietly = TRUE)) {
  library(Kendall)
  mk_test <- Kendall::MannKendall(us_wheat$`Wheat (tonnes per hectare)`)
  mk_table <- data.frame(
    Statistic = c("Tau", "p-value"),
    Value = c(
      round(mk_test$tau, 3),
      format.pval(mk_test$p.value, digits = 3)
    )
  )

  # Display the Mann-Kendall test results
  knitr::kable(mk_table,
               align = c("l", "r"),
               format = "html") %>%
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                             full_width = FALSE,
                             position = "center")
} else {
  message("The Kendall package is not installed. Install it with install.packages('Kendall') to run the Mann-Kendall trend test.")
}
```

::: {.callout-note}
## What the code is doing

`Kendall::MannKendall()` walks every pair of observations in the series and counts how often a later value exceeds an earlier one. Tau is that count, scaled to fall between -1 and 1.
:::

::: {.callout-important}
## Reading the results
Tau = 0.87 is about as strong a positive trend signal as you can get from this test, and p < 0.001 confirms it's not noise. The reason to like Mann-Kendall over a regression slope here is that it makes almost no assumptions: no normality, no linear form, no homoscedasticity. All it asks is whether the series is monotonic, and it answers based on rank ordering rather than magnitude.

That makes it the right tool when you suspect outliers, structural breaks, or a non-linear shape that would mislead `lm()`. The downside is that tau is unitless: it tells you the trend is there and how consistent it is, but not how big in tonnes per hectare per year. Pair it with the regression slope when you need both the "is there a trend" and "how much" answers in the same report.
:::

## Summary

This chapter has demonstrated a variety of statistical tests using real agricultural and biodiversity datasets. We've covered:

1. **Tests for comparing groups**:
   - t-tests for comparing two groups
   - ANOVA for comparing multiple groups
   - Non-parametric alternatives when data doesn't meet parametric assumptions

2. **Tests for relationships**:
   - Correlation analysis to measure the strength of relationships
   - Regression analysis to model relationships between variables

3. **Tests for categorical data**:
   - Chi-Square test for examining associations between categorical variables

4. **Tests for time series data**:
   - Time series analysis for identifying patterns over time
   - Mann-Kendall test for detecting trends

When conducting statistical tests, remember to:
- Clearly define your research question
- Check if your data meets the assumptions of the test
- Choose the appropriate test based on your data type and research question
- Interpret results in the context of your research question
- Consider the practical significance, not just statistical significance

## Chapter Summary

### Key Concepts

-   **Test Selection**: Choosing the right test depends on data type, distribution, and study design
-   **Parametric vs. Non-Parametric**: Parametric tests have more power but require normality; non-parametric tests are more robust
-   **Group Comparisons**: t-tests and ANOVA compare means across two or more groups
-   **Relationships**: Correlation and regression quantify associations between variables
-   **Post-hoc Testing**: Essential for ANOVA and Kruskal-Wallis to pinpoint specific differences

### R Functions Learned

-   `t.test()` - One-sample, two-sample, and paired t-tests
-   `aov()` - Analysis of Variance (ANOVA)
-   `TukeyHSD()` - Post-hoc test for ANOVA
-   `wilcox.test()` - Mann-Whitney U and Wilcoxon signed-rank tests
-   `kruskal.test()` - Kruskal-Wallis test
-   `cor.test()` - Correlation analysis (Pearson and Spearman)
-   `lm()` - Linear and multiple regression models
-   `shapiro.test()` - Test for normality

### Next Steps

In the next part of the book, we will shift our focus to **Data Visualization**, exploring how to create compelling and informative graphics to communicate your results effectively.

## Exercises

### Conceptual

1.  **Test Selection**: For each of the following scenarios, identify the appropriate statistical test:
    -   Comparing tree heights between three different soil types (normal distribution)
    -   Testing if the number of bird species differs between two forest fragments (non-normal)
    -   Predicting biomass based on rainfall and temperature

### Applied (using book datasets)

2.  **ANOVA vs. Kruskal-Wallis**: Using `../data/agriculture/crop_yields.csv`, compare maize yields between continents using both ANOVA and the Kruskal-Wallis test. Which is more appropriate, and why? Use plots and residual diagnostics; treat Shapiro-Wilk and Levene's test as supporting evidence rather than automatic pass/fail gates.

3.  **Correlation comparison**: Examine the relationship between potato and rice yields. Calculate both Pearson and Spearman correlations and explain which is more appropriate given the data's distribution.

4.  **Categorical associations**: Using `../data/ecology/biodiversity.csv`, investigate whether there's an association between conservation status and another categorical variable of your choice using a Chi-Square test.

5.  **Time series**: Perform a Mann-Kendall trend test on wheat yields for China and compare the trend slope with that of the United States.

6.  **Two-group comparison**: Using `../data/entomology/insects.csv` (animal-shelter outcomes, not insect measurements), choose a numeric state-count variable and compare two outcome groups or years. Explain the observational unit and why the test is appropriate.

7.  **Multiple regression**: Using `../data/economics/economic.csv`, fit a multiple regression model to predict coffee quality scores and interpret the slope, R-squared value, and residual diagnostics.

### Built-in datasets

8.  **Group Comparison**: Use the `iris` dataset to compare sepal lengths among the three species using ANOVA. Follow up with a post-hoc Tukey HSD test.

9.  **Non-Parametric Test**: Use the `mtcars` dataset to compare fuel efficiency (`mpg`) between automatic and manual transmissions (`am`) using a Mann-Whitney U test.

### Reporting

10. **Reporting**: Write a short paragraph reporting the results of one of your analyses above, including the test statistic, degrees of freedom, p-value, effect size, and a confidence interval where appropriate.
