---
prefer-html: true
---

# Regression Analysis

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

1.  Fit and interpret linear and multiple regression models in R
2.  Check regression assumptions using diagnostic plots
3.  Implement logistic regression for binary classification
4.  Apply the `tidymodels` framework for machine learning workflows
5.  Perform cross-validation to estimate model performance
6.  Tune hyperparameters to optimize model accuracy
:::

## Introduction

Regression is how we describe a relationship between variables with an equation we can poke at. The interpretation depends on the design: an observational regression coefficient is generally an adjusted association, not automatically a causal effect [@zuur2009mixed; @bolker2009generalized]. How much does tree height change per millimetre of rainfall? Does adding fertiliser actually increase yield once you account for soil type? This chapter covers the workhorse: linear regression with one predictor, multiple linear regression when you have several, and logistic regression when the outcome is yes-or-no instead of a number. We will also do the diagnostic work that separates a model worth quoting from one that just happens to fit your sample.

## Linear regression

Linear regression draws the best straight line through your data, where "best" means it minimises the squared vertical distances between the line and each point. You get an intercept, a slope, and an equation you can use to predict the response for new values of the predictor.

```{r}
#| label: setup-linear-regression
#| include: false

# Load required packages
library(tidyverse)      # For data manipulation
library(ggplot2)        # For visualization
library(broom)          # For tidying model outputs
library(performance)    # For model diagnostics
library(see)            # For visualization of model diagnostics
library(parameters)     # For parameter description
library(ggeffects)      # For visualizing model effects
library(gt)             # For tables

# Set a professional theme for all plots
theme_set(theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "gray90", fill = NA, linewidth = 0.5)
  ))

# Load the Palmer penguins dataset (stored as climate_data.csv)
penguins <- read_csv("../data/environmental/climate_data.csv", show_col_types = FALSE)

# Remove rows with missing values in the key variables we'll use for regression
penguins_clean <- penguins %>%
  filter(!is.na(bill_length_mm), !is.na(body_mass_g),
         !is.na(bill_depth_mm), !is.na(flipper_length_mm)) %>%
  # Add species as a factor for proper modeling
  mutate(species = as.factor(species))

# Create a linear regression model
model <- lm(body_mass_g ~ bill_length_mm, data = penguins_clean)

# Get model summary with broom for cleaner output
model_summary <- summary(model)
model_tidy <- tidy(model, conf.int = TRUE)
model_glance <- glance(model)
```

```{r}
#| label: model-output-linear
#| echo: false
# Display key model metrics
cat("Model Summary:\n")
cat(paste0("R² = ", round(model_glance$r.squared, 3),
          ", Adjusted R² = ", round(model_glance$adj.r.squared, 3), "\n"))
cat(paste0("F-statistic: ", round(model_glance$statistic, 2),
          " on ", model_glance$df, " and ", model_glance$df.residual,
          " DF, p-value: ", format.pval(model_glance$p.value, digits = 3), "\n\n"))
```

```{r}
#| label: tbl-linear-regression-coef
#| tbl-cap: "Linear Regression Coefficients"

# Create a more professional table of coefficients
model_tidy %>%
  gt() %>%
  tab_header(title = "Linear Regression Coefficients") %>%
  fmt_number(columns = c("estimate", "std.error", "statistic", "p.value", "conf.low", "conf.high"), decimals = 3)
```

```{r}
#| label: fig-linear-regression-plot
#| fig-cap: "Relationship between bill length and body mass"
#| fig-alt: "Scatter plot of penguin bill length (mm) vs body mass (g) with points colored by species (viridis), a dark-red linear regression line with pink CI band, and annotations showing R-squared, p-value, and the fitted equation."

# Create an enhanced scatter plot with regression line
ggplot(penguins_clean, aes(x = bill_length_mm, y = body_mass_g)) +
  # Add data points with some transparency
  geom_point(aes(color = species), alpha = 0.7, size = 3) +
  # Add regression line with confidence interval
  geom_smooth(method = "lm", color = "darkred", fill = "pink", alpha = 0.2) +
  # Add annotations for R² and p-value
  annotate("text", x = min(penguins_clean$bill_length_mm) + 1,
           y = max(penguins_clean$body_mass_g) - 500,
           label = paste0("R² = ", round(model_glance$r.squared, 3),
                         "\np < ", format.pval(model_glance$p.value, digits = 3)),
           hjust = 0, size = 4, color = "darkred") +
  # Add regression equation
  annotate("text", x = min(penguins_clean$bill_length_mm) + 1,
           y = max(penguins_clean$body_mass_g) - 1000,
           label = paste0("y = ", round(coef(model)[1], 1), " + ",
                         round(coef(model)[2], 1), "x"),
           hjust = 0, size = 4, color = "darkred") +
  # Add professional labels
  labs(
    title = "Relationship Between Bill Length and Body Mass",
    subtitle = "Linear regression analysis shows positive correlation with species differences",
    x = "Bill Length (mm)",
    y = "Body Mass (g)",
    color = "Species",
    caption = "Data source: Palmer Penguins dataset"
  ) +
  # Use a colorblind-friendly palette
  scale_color_viridis_d() +
  # Adjust axis limits for better visualization
  coord_cartesian(expand = TRUE)
```

```{r}
#| label: fig-linear-regression-diag
#| fig-cap: "Model diagnostic plots"
#| fig-alt: "Multi-panel diagnostic display from the performance package's check_model: posterior predictive check, linearity, homogeneity of variance, influential observations, collinearity, and normality of residuals for the bill-length regression."

# Create diagnostic plots using the performance package
check_model <- check_model(model)
plot(check_model)
```

::: {.callout-important}
## Reading the diagnostic plots

- **Posterior predictive check**: simulated draws should overlay the
  observed density; large gaps indicate model misspecification.
- **Linearity**: the reference line should be roughly flat; curvature
  means the linear form is wrong.
- **Homogeneity of variance**: points should scatter evenly around the
  line. A funnel shape signals heteroscedasticity; consider a
  transformation or robust standard errors.
- **Influential observations**: flagged points outside the contour
  exert disproportionate leverage. Refit without them to confirm the
  conclusion is not driven by 1-2 cases.
- **Collinearity (multi-predictor models)**: VIF > 5 warrants concern;
  > 10 means one predictor is largely redundant.
- **Normality of residuals**: Q-Q points should fall on the diagonal.
  Mild departures are fine at n > 30 (CLT); heavy tails matter for
  prediction intervals more than for slope inference.
:::

::: {.callout-note}
## What the code is doing

`lm(body_mass_g ~ bill_length_mm, data = penguins_clean)` fits the regression. The formula reads as "body mass explained by bill length". The scatterplot adds the fitted line with `geom_smooth(method = "lm")` and annotates R² and the equation directly on the figure so a reader can see both at once. The diagnostic panel comes from `performance::check_model()`, which bundles the residual, Q-Q, scale-location, and leverage plots into a single call.
:::

::: {.callout-important}
## Reading the results

See @fig-linear-regression-plot, @fig-linear-regression-diag, and @tbl-linear-regression-coef.

Start with the headline numbers. R² tells you what fraction of the variation in body mass the bill-length slope accounts for, and the p-value on the slope tells you whether that relationship is bigger than sampling noise. A positive slope means heavier penguins tend to have longer bills; the size of the slope is how many grams of body mass you get per extra millimetre of bill.

Then look at whether the model has earned the right to make those claims. The diagnostic panel is where I check that: residuals scattered around zero with no funnel, Q-Q points roughly on the diagonal, no single point dragging the line around. If any of those go sideways, the coefficient is still a number but I would not bet on the standard error.

Finally, ask whether the effect is biologically interesting, not just statistically detectable. With a few hundred penguins you can pick up tiny slopes that mean nothing in the field. The confidence interval on the slope and the R² together are usually the honest summary.
:::

::: {.callout-tip}
## A working checklist for regression

Before fitting, look at the data. Plot the predictor against the response. If the cloud is curved, a straight line is the wrong tool. If there are obvious outliers, decide whether they belong in the analysis before they decide the slope for you.

After fitting, work through the diagnostics in order: residuals first, then leverage, then VIF if you have more than one predictor. Refit without any influential cases and see whether your conclusion changes. If it does, that is part of the result.

When you write it up, give the coefficient, its confidence interval, the sample size, and at least one diagnostic. A p-value alone is not a report.
:::

## Multiple regression

Most ecological questions have more than one moving part. Body mass is not just about bill length; it also depends on flipper length, on bill depth, and on which species you are looking at. Multiple regression lets you put several predictors on the right-hand side of the equation and ask what each one contributes once the others are held constant.

```{r}
#| label: setup-multiple-regression
#| include: false

# Load required packages
library(broom)
library(knitr)
library(dplyr)
library(car)
library(effects)

# Create multiple regression model
multi_model <- lm(body_mass_g ~ bill_length_mm + bill_depth_mm + flipper_length_mm + species,
                 data = penguins_clean)

# Get model summary with broom for cleaner output
multi_summary <- summary(multi_model)
multi_tidy <- tidy(multi_model, conf.int = TRUE)
multi_glance <- glance(multi_model)
```

```{r}
#| label: model-output-multiple
#| echo: false
# Display key model metrics
cat("Multiple Regression Model Summary:\n")
cat(paste0("R² = ", round(multi_glance$r.squared, 3),
          ", Adjusted R² = ", round(multi_glance$adj.r.squared, 3), "\n"))
cat(paste0("F-statistic: ", round(multi_glance$statistic, 2),
          " on ", multi_glance$df, " and ", multi_glance$df.residual,
          " DF, p-value: ", format.pval(multi_glance$p.value, digits = 3), "\n\n"))

# Create a simple table of coefficients
knitr::kable(multi_tidy, digits = 3, caption = "Multiple Regression Coefficients")
```

```{r}
#| label: tbl-multiple-regression-vif
#| tbl-cap: "Variance Inflation Factors (VIF)"

# Check for multicollinearity
vif_values <- car::vif(multi_model)
knitr::kable(vif_values, digits = 3)
```

```{r}
#| label: fig-multiple-regression-diag
#| fig-cap: "Multiple regression diagnostic plots"
#| fig-alt: "performance::check_model panel grid for the multiple regression of body mass on bill length, bill depth, flipper length, and species, including posterior predictive check, linearity, residual variance, influential points, collinearity, and Q-Q plot."

# Check for model assumptions
check_multi_model <- check_model(multi_model)
plot(check_multi_model)
```

```{r}
#| label: fig-multiple-regression-effects
#| fig-cap: "Predictor effects"
#| fig-alt: "effects::allEffects panel of partial-effect plots for each predictor in the multiple regression (bill length, bill depth, flipper length, species), each showing the fitted relationship to body mass with confidence bands."

# Visualize predictor effects
plot(allEffects(multi_model), ask = FALSE)
```

```{r}
#| label: tbl-model-comparison
#| tbl-cap: "Comparison of Regression Models"

# Compare models using base R
model_comparison <- data.frame(
  Metric = c("R²", "Adjusted R²", "AIC", "BIC", "N"),
  `Simple Model` = c(
    round(summary(model)$r.squared, 3),
    round(summary(model)$adj.r.squared, 3),
    round(AIC(model), 1),
    round(BIC(model), 1),
    nobs(model)
  ),
  `Multiple Model` = c(
    round(summary(multi_model)$r.squared, 3),
    round(summary(multi_model)$adj.r.squared, 3),
    round(AIC(multi_model), 1),
    round(BIC(multi_model), 1),
    nobs(multi_model)
  ),
  check.names = FALSE
)
knitr::kable(model_comparison)
```

```{r}
#| label: fig-predicted-vs-actual
#| fig-cap: "Predicted vs actual body mass"
#| fig-alt: "Scatter plot of predicted body mass (x) vs observed body mass (y), points colored by penguin species with a dashed 1:1 reference line; points cluster tightly along the line, indicating good model fit."

# Create a visualization of predicted vs. actual values
predicted_values <- augment(multi_model, data = penguins_clean)

ggplot(predicted_values, aes(x = .fitted, y = body_mass_g, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  labs(
    title = "Predicted vs. Actual Body Mass",
    subtitle = "Points closer to the dashed line indicate better predictions",
    x = "Predicted Body Mass (g)",
    y = "Actual Body Mass (g)",
    color = "Species",
    caption = "Model: body_mass_g ~ bill_length_mm + bill_depth_mm + flipper_length_mm + species"
  ) +
  scale_color_viridis_d() +
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-flipper-pdp
#| fig-cap: "Effect of flipper length on body mass"
#| fig-alt: "ggeffects partial-dependence line plot of predicted body mass (g) as a function of flipper length (mm), holding other predictors constant; the trend is positive and roughly linear with a 95% confidence band."

# Create a partial dependence plot for flipper length
pdp_flipper <- ggeffects::ggpredict(multi_model, terms = "flipper_length_mm")
plot(pdp_flipper) +
  labs(
    title = "Effect of Flipper Length on Body Mass",
    subtitle = "Controlling for other variables in the model",
    caption = "Shaded area represents 95% confidence interval"
  )
```

::: {.callout-note}
## What the code is doing

The formula `body_mass_g ~ bill_length_mm + bill_depth_mm + flipper_length_mm + species` puts four predictors in the model. R handles species as a categorical variable automatically, encoding it as dummy contrasts. `car::vif()` reports variance inflation factors so you can see whether the morphometric predictors are stepping on each other. `effects::allEffects()` then draws the partial relationship for each predictor with the others held at their means, which is what you actually want to see when there are several variables in play. `ggeffects::ggpredict()` does the same job for one predictor at a time and gives you a cleaner ggplot object to work with.
:::

::: {.callout-important}
## Reading the results

See @tbl-multiple-regression-vif, @fig-multiple-regression-diag, @fig-multiple-regression-effects, and @tbl-model-comparison.

Adjusted R² jumps once flipper length and species enter the model, which tells you bill length alone was leaving a lot on the table. Flipper length and bill length both have positive slopes: heavier birds are longer everywhere. Bill depth drops to a weaker contribution once you control for the other measurements, which is the typical sign that it was a partial proxy for one of them in the simple model. The species coefficients pick up the average body-mass offset between Gentoo, Chinstrap, and Adelie that morphometry alone cannot capture.

VIFs are all comfortably under 5, so collinearity is not distorting the standard errors. The diagnostic panel looks clean: no funnel in the residuals, Q-Q points on the line, no influential cases sitting outside the contours. That is the regression behaving itself.
:::

::: {.callout-tip}
## A working checklist for multiple regression

Pick predictors that have a reason to be there. Throwing every column into `lm()` and letting AIC sort it out is a recipe for a model that fits this dataset and nothing else.

Check correlations among predictors before you fit. If two of them are above about 0.8, decide which one stays. VIFs over 10 are a hard stop; over 5 is a yellow flag worth investigating.

Report effects, not just p-values. A coefficient with its confidence interval, alongside a partial effect plot, communicates more in two seconds than a row of asterisks ever will. Standardised coefficients are useful when predictors are on different scales and you want a fair comparison of their relative importance.
:::

## Logistic regression

Sometimes the outcome you care about is not a number but a yes-or-no. Did the seed germinate? Is this individual an Adelie or something else? Linear regression is the wrong tool for that, because nothing constrains its predictions to fall between 0 and 1. Logistic regression solves that by modelling the log-odds of the outcome as a linear function of the predictors, then squashing the result back into a probability through the logit link.

```{r}
#| label: setup-logistic-regression
#| include: false

# Load required packages
library(pROC)
library(DHARMa)

# Prepare data for logistic regression by creating a binary outcome
# We'll predict whether a penguin is Adelie species or not
penguins_binary <- penguins_clean %>%
  # Create a binary outcome variable (is_adelie)
  mutate(is_adelie = ifelse(species == "Adelie", 1, 0),
         # Convert to factor for better model interpretation
         is_adelie_factor = factor(is_adelie, levels = c(0, 1),
                                  labels = c("Other", "Adelie")))

# Create logistic regression model
log_model <- glm(is_adelie ~ bill_depth_mm + body_mass_g,
                family = binomial(link = "logit"),
                data = penguins_binary)

# Summarize model with tidy output
log_summary <- summary(log_model)
log_tidy <- tidy(log_model, conf.int = TRUE, exponentiate = TRUE)
log_glance <- glance(log_model)
```

```{r}
#| label: model-output-logistic
#| echo: false
# Display key model metrics
cat("Logistic Regression Model Summary:\n")
cat(paste0("AIC: ", round(log_glance$AIC, 2),
          ", Deviance: ", round(log_glance$deviance, 2), "\n"))
cat(paste0("Null Deviance: ", round(log_glance$null.deviance, 2),
          ", DF: ", log_glance$df.null, ", Residual DF: ", log_glance$df.residual, "\n\n"))

# Calculate and show pseudo R-squared (McFadden)
pseudo_r2 <- 1 - (log_glance$deviance / log_glance$null.deviance)
cat(paste0("McFadden's Pseudo R²: ", round(pseudo_r2, 3), "\n\n"))
```

```{r}
#| label: tbl-logistic-odds
#| tbl-cap: "Logistic Regression Results (Odds Ratios)"

# Create a table of odds ratios with CI
log_tidy %>%
  gt() %>%
  tab_header(title = "Logistic Regression Results (Odds Ratios)") %>%
  fmt_number(columns = c("estimate", "std.error", "statistic", "p.value", "conf.low", "conf.high"), decimals = 3)
```

```{r}
#| label: tbl-logistic-confusion
#| tbl-cap: "Confusion Matrix"

# Add predictions to the data
penguins_pred <- penguins_binary %>%
  mutate(
    predicted_prob = predict(log_model, type = "response"),
    predicted_class = ifelse(predicted_prob > 0.5, "Adelie", "Other"),
    correct = ifelse(predicted_class == ifelse(is_adelie == 1, "Adelie", "Other"), "Correct", "Incorrect")
  )

# Create a confusion matrix
confusion <- table(
  Predicted = penguins_pred$predicted_class,
  Actual = ifelse(penguins_pred$is_adelie == 1, "Adelie", "Other")
)

# Calculate accuracy metrics
accuracy <- sum(diag(confusion)) / sum(confusion)
sensitivity <- confusion["Adelie", "Adelie"] / sum(confusion[, "Adelie"])
specificity <- confusion["Other", "Other"] / sum(confusion[, "Other"])
precision <- confusion["Adelie", "Adelie"] / sum(confusion["Adelie", ])

# Display confusion matrix with metrics
knitr::kable(confusion,
            caption = paste0("Confusion Matrix (Accuracy = ", round(accuracy * 100, 1), "%)"))
```

```{r}
#| label: model-metrics-logistic
#| echo: false
cat(paste0("Sensitivity (True Positive Rate): ", round(sensitivity * 100, 1), "%\n"))
cat(paste0("Specificity (True Negative Rate): ", round(specificity * 100, 1), "%\n"))
cat(paste0("Precision (Positive Predictive Value): ", round(precision * 100, 1), "%\n\n"))
```

```{r}
#| label: fig-logistic-roc
#| fig-cap: "ROC curve for Adelie penguin classification"
#| fig-alt: "Receiver operating characteristic curve from pROC for the logistic model predicting Adelie vs other species; the curve hugs the upper-left corner and the AUC value is printed inside the shaded area under the curve."

# Create ROC curve
roc_curve <- roc(penguins_binary$is_adelie, fitted(log_model))
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, print.auc = TRUE, auc.polygon = TRUE,
     grid = TRUE, main = "ROC Curve for Adelie Penguin Classification")
```

```{r}
#| label: fig-logistic-probs
#| fig-cap: "Classification visualization by species"
#| fig-alt: "Faceted scatter plot (one panel per species) of bill depth (mm) vs body mass (g) with points colored by predicted Adelie probability on a navy-white-red gradient; a black p=0.5 contour shows the decision boundary and ✗ marks misclassified points."

# Create a visualization of predicted probabilities by species
ggplot(penguins_pred, aes(x = bill_depth_mm, y = body_mass_g, color = predicted_prob)) +
  geom_point(size = 3, alpha = 0.7) +
  # Add decision boundary (0.5 probability contour)
  geom_contour(aes(z = predicted_prob), breaks = 0.5, color = "black", linewidth = 1) +
  # Add text labels for misclassified points
  geom_text(data = filter(penguins_pred, correct == "Incorrect"),
            aes(label = "✗"), color = "black", size = 4, nudge_y = 0.5) +
  # Color gradient for probability
  scale_color_gradient2(low = "navy", mid = "white", high = "red",
                      midpoint = 0.5, limits = c(0, 1)) +
  facet_wrap(~species) +
  labs(
    title = "Classification of Adelie vs. Other Penguins",
    subtitle = paste0("AUC = ", round(auc_value, 3), ", Accuracy = ", round(accuracy * 100, 1), "%"),
    x = "Bill Depth (mm)",
    y = "Body Mass (g)",
    color = "Probability\nof Adelie",
    caption = "Black line: decision boundary (p=0.5), ✗: misclassified points"
  ) +
  theme(legend.position = "right")
```

```{r}
#| label: fig-logistic-effects
#| fig-cap: "Marginal effects on probability"
#| fig-alt: "effects::allEffects panel of marginal-effect curves showing each logistic-regression predictor's influence on the probability of being Adelie, with 95% confidence bands."

# Create marginal effects plots for the predictors
plot(allEffects(log_model), ask = FALSE, main = "Marginal Effects on Probability of Adelie")
```

```{r}
#| label: fig-logistic-residuals
#| fig-cap: "DHARMa residual diagnostics"
#| fig-alt: "DHARMa two-panel diagnostic plot for the logistic regression: a Q-Q plot of simulated residuals with uniformity and dispersion tests on the left, and a residuals vs predicted plot with quantile regression lines on the right."

# Check model fit
sim_residuals <- simulateResiduals(log_model)
plot(sim_residuals)
```

::: {.callout-note}
## What the code is doing

`glm(is_adelie ~ bill_depth_mm + body_mass_g, family = binomial(link = "logit"))` fits the logistic model. The `tidy(..., exponentiate = TRUE)` call converts coefficients into odds ratios, which are easier to talk about than log-odds. McFadden's pseudo-R² is computed by hand from the deviances since `glance()` does not report it directly. The confusion matrix uses a 0.5 probability cutoff to convert fitted probabilities into class labels, and `pROC::roc()` builds the ROC curve from the underlying probabilities so you do not have to commit to any single threshold. `DHARMa::simulateResiduals()` runs simulation-based residual checks, which are far more reliable for GLMs than the raw Pearson or deviance residuals.
:::

::: {.callout-important}
## Reading the results

The AUC is the headline number for classification. It is the probability that a randomly chosen Adelie gets a higher predicted score than a randomly chosen non-Adelie, and values above about 0.9 mean the model separates the two groups cleanly. Accuracy, sensitivity, and specificity flesh that out at the 0.5 cutoff, but you can read off other operating points from the ROC curve if the cost of false positives and false negatives is not equal.

The odds ratios tell you how a one-unit change in each predictor multiplies the odds of being Adelie. An odds ratio of 2 doubles the odds; 0.5 halves them. Bill depth comes out as a strong discriminator, which lines up with what is known about Adelie morphology relative to Gentoo and Chinstrap.

The misclassified points are worth looking at, not glossing over. They tend to sit in the morphological overlap between species, and they are often the most informative birds in the dataset. The DHARMa residual plot is the sanity check: if the simulated residuals look uniform and the dispersion test passes, the binomial assumption is holding up.
:::

::: {.callout-tip}
## A working checklist for logistic regression

Mind the events-per-predictor ratio. The usual rule of thumb is at least ten events of the rarer class for each predictor in the model. With fewer, your coefficients will be unstable no matter what the p-values say.

Pick a threshold based on what the errors cost, not by default. A 0.5 cutoff makes sense when false positives and false negatives are equally bad. They rarely are.

Report odds ratios with confidence intervals, an AUC, and at least one calibration check. Coefficients on the logit scale are technically correct but almost nobody reads them at a glance.
:::

## Summary so far

So far we have done the base-R version of regression. `lm()` handles continuous outcomes, with one predictor or several, and gives you coefficients, confidence intervals, and a diagnostic panel you can read in about a minute. `glm()` extends the same machinery to binary outcomes through the logit link, with odds ratios and an ROC curve as the natural way to report what came out.

The next section does the same modelling with tidymodels. The maths is identical. What changes is the scaffolding around it: explicit train/test splits, a recipe for preprocessing, a workflow object that bundles model and recipe together, and built-in support for resampling and tuning. For one quick `lm()` call it is overkill. For comparing several models, or for anything you plan to deploy or revisit, it pays for itself fast.

## Inference and prediction are different goals

The regression examples above estimate associations and explain how a model relates predictors to an outcome. The tidymodels workflow below is primarily for prediction. A useful predictive score does not establish causation, and a statistically precise coefficient does not guarantee accurate predictions for new data. State the goal before choosing the workflow.

For prediction, keep all preprocessing, feature selection, and tuning inside the resampling workflow to avoid leakage. Random folds are inappropriate when future observations, sites, plots, or species must be predicted: use temporal, spatial, or grouped resampling that mirrors deployment. Report a baseline model, uncertainty, calibration, and metrics suited to the outcome; do not rely on accuracy alone for imbalanced classes.

::: {.callout-warning}
A rule such as “ten events per predictor” is only a rough historical heuristic, not a universal sample-size guarantee. Check for separation, sparse data, unstable estimates, calibration, and out-of-sample performance. Consider penalized models or simpler predictors when information is limited.
:::

## Regression with the tidymodels framework

Tidymodels is a set of packages that share a common interface for modelling. You define preprocessing in a recipe, the model in a parsnip spec, and tie them together with a workflow. Once that is in place, swapping a random forest in for a linear model is a one-line change, and cross-validation, tuning, and metrics all work the same way regardless of which engine is under the hood.

### Setting Up Tidymodels

```{r}
#| label: tidymodels-setup

# Load tidymodels packages
library(tidymodels)  # Meta-package loading parsnip, recipes, rsample, tune, workflows, yardstick
library(tidyverse)    # For data manipulation

# Set random seed for reproducibility
set.seed(123)

# Load and prepare data
penguins <- read_csv("../data/environmental/climate_data.csv", show_col_types = FALSE) %>%
  filter(!is.na(bill_length_mm), !is.na(body_mass_g),
         !is.na(bill_depth_mm), !is.na(flipper_length_mm)) %>%
  mutate(species = as.factor(species))

# Display data structure
glimpse(penguins)
```

::: {.callout-note}
## What tidymodels is trying to do

The framework leans on a few habits. The same syntax fits a linear model and a random forest, so switching engines does not mean relearning the API. Preprocessing, model fitting, resampling, and evaluation each live in their own package, so you can mix and match. Everything plays nicely with dplyr and ggplot2 pipelines. Resampling and validation are first-class rather than something you bolt on afterwards. And adding a new model type is a matter of writing a parsnip wrapper, not rewiring the whole stack.
:::

### Linear Regression with Tidymodels

```{r}
#| label: tidymodels-linear-setup
#| echo: true

# Step 1: Split data into training and testing sets
penguin_split <- initial_split(penguins, prop = 0.75, strata = species)
penguin_train <- training(penguin_split)
penguin_test <- testing(penguin_split)

cat("Training set:", nrow(penguin_train), "observations\n")
cat("Testing set:", nrow(penguin_test), "observations\n")

# Step 2: Define a recipe for preprocessing
penguin_recipe <- recipe(body_mass_g ~ bill_length_mm + bill_depth_mm +
                         flipper_length_mm + species,
                         data = penguin_train) %>%
  step_dummy(species) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_zv(all_predictors())

# Step 3: Specify the model
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

# Step 4: Create a workflow combining recipe and model
penguin_wf <- workflow() %>%
  add_recipe(penguin_recipe) %>%
  add_model(lm_spec)

# Step 5: Fit the model
penguin_fit <- penguin_wf %>%
  fit(data = penguin_train)
```

```{r}
#| label: tbl-tidymodels-coef
#| tbl-cap: "Tidymodels Linear Regression Coefficients"

tidy(penguin_fit) %>%
  knitr::kable(digits = 3)
```

```{r}
#| label: tbl-tidymodels-train
#| tbl-cap: "Training Set Performance Metrics"

train_results <- penguin_fit %>%
  predict(penguin_train) %>%
  bind_cols(penguin_train) %>%
  metrics(truth = body_mass_g, estimate = .pred)

knitr::kable(train_results, digits = 3)
```

```{r}
#| label: tbl-tidymodels-test
#| tbl-cap: "Test Set Performance Metrics"

test_results <- penguin_fit %>%
  predict(penguin_test) %>%
  bind_cols(penguin_test) %>%
  metrics(truth = body_mass_g, estimate = .pred)

knitr::kable(test_results, digits = 3)
```

```{r}
#| label: fig-tidymodels-linear
#| fig-cap: "Tidymodels predicted vs actual body mass"
#| fig-alt: "Test-set scatter plot of actual body mass (x) vs predicted body mass (y) from the tidymodels linear workflow, with viridis-colored species and a dashed 1:1 reference line indicating predictions track observed values closely."

penguin_pred <- penguin_fit %>%
  predict(penguin_test) %>%
  bind_cols(penguin_test)

ggplot(penguin_pred, aes(x = body_mass_g, y = .pred, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_viridis_d() +
  labs(
    title = "Tidymodels: Predicted vs. Actual Body Mass",
    subtitle = "Test set predictions from linear regression workflow",
    x = "Actual Body Mass (g)",
    y = "Predicted Body Mass (g)",
    color = "Species",
    caption = "Dashed line represents perfect predictions"
  ) +
  theme_minimal()
```

::: {.callout-important}
## Why this is worth the extra typing

The workflow forces you to split the data, preprocess it, fit the model, and evaluate it as separate steps. That sounds bureaucratic but it pays off. Train and test stay properly separated, so you cannot accidentally peek. The recipe encodes preprocessing once and applies it consistently to new data, which is the part people most often get wrong by hand. The workflow object is reusable: you can drop new data into the same pipeline without rewriting it. And metrics are computed the same way no matter which model you fit, so model comparisons are fair.

The test-set RMSE and R² are the numbers to report. They are honest estimates of how the model will do on data it has not seen.
:::

### Cross-Validation with Tidymodels

```{r}
#| label: cross-validation-setup
#| echo: true

# Create cross-validation folds
set.seed(456)
penguin_folds <- vfold_cv(penguin_train, v = 10, strata = species)

cat("Created", nrow(penguin_folds), "cross-validation folds\n")

# Fit model using cross-validation
cv_results <- penguin_wf %>%
  fit_resamples(
    resamples = penguin_folds,
    control = control_resamples(save_pred = TRUE),
    metrics = metric_set(rmse, rsq, mae)
  )
```

```{r}
#| label: tbl-cv-metrics
#| tbl-cap: "10-Fold Cross-Validation Performance"

cv_metrics <- collect_metrics(cv_results)
knitr::kable(cv_metrics, digits = 3)
```

```{r}
#| label: fig-cv-distribution
#| fig-cap: "Cross-validation performance distribution"
#| fig-alt: "Boxplot with jittered points of per-fold metric values (RMSE, R-squared, MAE) from 10-fold cross-validation; tight box widths indicate consistent performance across folds."

collect_metrics(cv_results, summarize = FALSE) %>%
  ggplot(aes(x = .metric, y = .estimate)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  labs(
    title = "Cross-Validation Performance Distribution",
    subtitle = "10-fold CV shows model stability across folds",
    x = "Metric",
    y = "Value"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-cv-predictions
#| fig-cap: "Cross-validation predictions"
#| fig-alt: "Scatter plot of actual vs cross-validated predicted body mass for all out-of-fold predictions in steel blue, with a red dashed 1:1 line; points hug the line, confirming generalization across folds."

cv_predictions <- collect_predictions(cv_results)

ggplot(cv_predictions, aes(x = body_mass_g, y = .pred)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Cross-Validation Predictions",
    subtitle = "All predictions from 10-fold CV",
    x = "Actual Body Mass (g)",
    y = "Predicted Body Mass (g)"
  ) +
  theme_minimal()
```

::: {.callout-tip}
## Why bother with cross-validation

A single train/test split gives you one number. Cross-validation gives you ten, and the spread across folds tells you how much that single number could have wobbled. Every observation gets used for both training and validation across the folds, so you are not throwing data away. A big gap between training and CV metrics is the clearest signal of overfitting you can ask for. And when you are comparing models or tuning hyperparameters, CV is the only fair basis: anything else is comparing apples to whatever happened to be in your one test split.
:::

### Model Comparison with Tidymodels

```{r}
#| label: model-comparison-setup
#| echo: true

# Define multiple model specifications
lm_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

rf_spec <- rand_forest(trees = 100) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Create workflows for each model
lm_wf <- workflow() %>%
  add_recipe(penguin_recipe) %>%
  add_model(lm_spec)

rf_wf <- workflow() %>%
  add_recipe(penguin_recipe) %>%
  add_model(rf_spec)

# Fit both models with cross-validation
set.seed(789)
lm_cv <- lm_wf %>%
  fit_resamples(
    resamples = penguin_folds,
    metrics = metric_set(rmse, rsq, mae)
  )

rf_cv <- rf_wf %>%
  fit_resamples(
    resamples = penguin_folds,
    metrics = metric_set(rmse, rsq, mae)
  )

# Compare model performance
model_comparison <- bind_rows(
  collect_metrics(lm_cv) %>% mutate(model = "Linear Regression"),
  collect_metrics(rf_cv) %>% mutate(model = "Random Forest")
)
```

```{r}
#| label: tbl-model-comparison-rf
#| tbl-cap: "Model Comparison: Linear Regression vs. Random Forest"

knitr::kable(model_comparison %>% select(model, .metric, mean, std_err),
             digits = 3,
             col.names = c("Model", "Metric", "Mean", "Std Error"))
```

```{r}
#| label: fig-model-comparison
#| fig-cap: "Model performance comparison"
#| fig-alt: "Faceted bar chart comparing linear regression vs random forest across cross-validation metrics (MAE, RMSE, R-squared), with viridis-filled bars and error bars showing standard error around each mean."

model_comparison %>%
  ggplot(aes(x = model, y = mean, fill = model)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                width = 0.2) +
  facet_wrap(~ .metric, scales = "free_y") +
  scale_fill_viridis_d(begin = 0.3, end = 0.8) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Cross-validation results with standard errors",
    x = "Model Type",
    y = "Performance (mean ± SE)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Hyperparameter tuning

Random forests and most other machine-learning models have settings that the model itself does not learn from the data. For a random forest, `mtry` (how many predictors to consider at each split) and `min_n` (the minimum node size) are the usual two. Tuning means trying a grid of values and picking whichever combination gives the best cross-validated performance.

```{r}
#| label: fig-hyperparameter-tuning
#| fig-cap: "Hyperparameter tuning results"
#| fig-alt: "Two-panel line-and-point plot of random forest cross-validated RMSE against the mtry and min_n tuning parameters; each facet shows how RMSE changes as one hyperparameter varies, highlighting the optimal region."

# Define a model specification with tunable parameters
rf_tune_spec <- rand_forest(
  trees = 1000,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine("ranger") %>%
  set_mode("regression")

# Create a workflow with the tunable model
rf_tune_wf <- workflow() %>%
  add_recipe(penguin_recipe) %>%
  add_model(rf_tune_spec)

# Define the grid of parameter values to try
rf_grid <- grid_regular(
  mtry(range = c(1, 5)),
  min_n(range = c(2, 10)),
  levels = 5
)

# Tune the model using cross-validation
# Note: This may take a moment to run
set.seed(345)
rf_tune_res <- tune_grid(
  rf_tune_wf,
  resamples = penguin_folds,
  grid = rf_grid,
  metrics = metric_set(rmse, rsq)
)

# Collect tuning metrics
rf_tune_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  pivot_longer(cols = c(mtry, min_n), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, y = mean, color = parameter)) +
  geom_point() +
  geom_line() +
  facet_wrap(~parameter, scales = "free_x") +
  labs(title = "Hyperparameter Tuning Results", y = "RMSE (lower is better)") +
  theme_minimal()

# Select the best parameters
best_rf <- select_best(rf_tune_res, metric = "rmse")
print(best_rf)

# Finalize the workflow with the best parameters
final_rf_wf <- finalize_workflow(rf_tune_wf, best_rf)

# Fit the final model to the full training set and evaluate on test set
final_fit <- last_fit(final_rf_wf, penguin_split)

# Show final performance
collect_metrics(final_fit)
```

::: {.callout-tip}
## Picking a tuning strategy

Grid search just tries everything on the grid. It is exhaustive and simple, and fine for two or three parameters. Once the parameter space gets bigger, `tune_bayes()` uses Bayesian optimisation to spend more time in the promising regions and less on the obvious losers, which is usually much faster.
:::

## Chapter summary

### What we covered

Linear regression models a continuous outcome as a straight-line function of one or more predictors. Logistic regression does the same for a binary outcome via the logit link. Tidymodels (parsnip, recipes, rsample, tune, yardstick) gives all of that a common interface. Cross-validation estimates how the model will perform on data it has not seen, and tuning chooses the settings that make it perform best. None of this is worth quoting if the diagnostics are bad: normality, equal variance, linearity, and independence still matter regardless of which framework you used to fit the model.

### Functions worth remembering

`lm()` and `glm()` are the base-R workhorses. In tidymodels, `linear_reg()`, `logistic_reg()`, and `rand_forest()` are the model specs; `recipe()` defines preprocessing; `vfold_cv()` builds folds; `fit_resamples()` runs CV; `tune_grid()` does grid search; and `collect_metrics()` pulls the numbers out at the end.

### What's next

The next chapter, [Advanced Modeling Techniques](09-advanced-modeling.qmd), looks at how to interpret black-box models when the coefficients are not available, and how to handle data that comes in over time.

### Logistic regression with tidymodels

```{r}
#| label: logistic-setup
#| echo: true

# Prepare binary classification data
penguins_binary <- penguins %>%
  mutate(is_adelie = factor(ifelse(species == "Adelie", "Adelie", "Other"),
                           levels = c("Other", "Adelie")))

# Split data
set.seed(2023)
binary_split <- initial_split(penguins_binary, prop = 0.75, strata = is_adelie)
binary_train <- training(binary_split)
binary_test <- testing(binary_split)

# Create recipe
logistic_recipe <- recipe(is_adelie ~ bill_depth_mm + body_mass_g,
                         data = binary_train) %>%
  step_normalize(all_numeric_predictors())

# Specify logistic regression model
logistic_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# Create workflow
logistic_wf <- workflow() %>%
  add_recipe(logistic_recipe) %>%
  add_model(logistic_spec)

# Fit the model
logistic_fit <- logistic_wf %>%
  fit(data = binary_train)

# Get predictions with probabilities
logistic_pred <- logistic_fit %>%
  predict(binary_test, type = "prob") %>%
  bind_cols(logistic_fit %>% predict(binary_test)) %>%
  bind_cols(binary_test)
```

```{r}
#| label: tbl-logistic-metrics
#| tbl-cap: "Logistic Regression Classification Metrics"

logistic_metrics <- logistic_pred %>%
  metrics(truth = is_adelie, estimate = .pred_class, .pred_Adelie)

knitr::kable(logistic_metrics, digits = 3)
```

```{r}
#| label: fig-logistic-confusion
#| fig-cap: "Confusion matrix for Adelie classification"
#| fig-alt: "2x2 confusion-matrix heatmap (yardstick autoplot) for the tidymodels logistic classifier; diagonal cells (correct predictions) are dark, off-diagonals are pale, summarizing classification performance on the test set."

conf_mat(logistic_pred, truth = is_adelie, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Confusion Matrix: Adelie Classification") +
  theme_minimal()
```

```{r}
#| label: fig-logistic-roc-rf
#| fig-cap: "ROC curve for tidymodels logistic regression"
#| fig-alt: "yardstick autoplot of the ROC curve for the tidymodels logistic Adelie classifier; the curve bows toward the top-left corner, indicating strong discrimination between Adelie and other penguins."

logistic_pred %>%
  roc_curve(truth = is_adelie, .pred_Adelie) %>%
  autoplot() +
  labs(title = "ROC Curve: Tidymodels Logistic Regression") +
  theme_minimal()

# Calculate AUC for use in next plot
auc_value <- logistic_pred %>%
  roc_auc(truth = is_adelie, .pred_Adelie) %>%
  pull(.estimate)

cat("Area Under the Curve (AUC):", round(auc_value, 3), "\n")
```

```{r}
#| label: fig-logistic-probs-rf
#| fig-cap: "Predicted probabilities for Adelie classification"
#| fig-alt: "Scatter of bill depth (mm) vs body mass (g) with point color encoding predicted P(Adelie) on a navy-white-red gradient and point shape encoding the true species label; Adelie observations cluster in the high-probability red region."

ggplot(logistic_pred, aes(x = bill_depth_mm, y = body_mass_g,
                         color = .pred_Adelie, shape = is_adelie)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_gradient2(low = "navy", mid = "white", high = "red",
                       midpoint = 0.5) +
  labs(
    title = "Predicted Probabilities: Adelie Classification",
    subtitle = paste0("AUC = ", round(auc_value, 3)),
    x = "Bill Depth (mm)",
    y = "Body Mass (g)",
    color = "P(Adelie)",
    shape = "Actual Species"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Tidymodels for classification

The classification side of tidymodels works the same way as the regression side. `metrics()` gives you accuracy and ROC AUC without having to write them by hand. `predict(type = "prob")` returns class probabilities directly, which is what you want for thresholding and ROC curves. `conf_mat()` and `roc_curve()` have built-in `autoplot()` methods, so the standard diagnostic plots are one call away. And the workflow object behaves the same whether you are predicting a number or a class.
:::

## Base R vs tidymodels: when to use which

Reach for base R when you want to fit a model quickly, when there are only one or two predictors, when you are teaching the underlying statistics, or when you need a specific diagnostic that the base packages provide more directly. The output is the kind of summary table that statisticians have been reading for fifty years.

Reach for tidymodels when you are comparing several models, when preprocessing has more than one step, when you need cross-validation or hyperparameter tuning, or when you expect to revisit the analysis later and want a pipeline that is easy to rerun. The upfront cost is real; the payoff comes when the project gets bigger than a single script.

### Feature comparison

| Feature | Base R | Tidymodels |
|---------|--------|------------
| **Learning Curve** | Moderate | Steeper initially |
| **Consistency** | Varies by package | Unified interface |
| **Preprocessing** | Manual | Automated with recipes |
| **Cross-validation** | Manual setup | Built-in support |
| **Model Comparison** | Manual | Streamlined |
| **Hyperparameter Tuning** | Package-specific | Unified with tune |
| **Production Deployment** | More complex | Workflow objects |
| **Extensibility** | High | Very high |

::: {.callout-tip}
## Using both, in practice

What I usually do: start in base R while I am still figuring out what the model should look like. `lm()` and `glm()` are fast to type and the diagnostics are right there. Once I have something I trust, I rebuild it as a tidymodels workflow so I can cross-validate it properly and swap engines without rewriting everything. For a report or paper, the base-R summary is often what the reviewers expect to see. For a pipeline that someone else will rerun in six months, tidymodels is the safer bet.
:::

## Exercises

### Base R Exercises

1. Fit a linear regression model predicting body mass from bill length and bill depth. Create a 3D visualization of this relationship using the `plotly` package.

2. Conduct multiple regression with interaction terms between predictors. How does adding interactions improve model performance?

3. Perform model selection using AIC to find the most parsimonious multiple regression model for the penguin data.

4. Create a multinomial logistic regression model to classify all three penguin species based on morphological traits.

5. Compare the performance of logistic regression with other classification methods (e.g., random forest, support vector machines) for species identification.

### Tidymodels Exercises

6. Create a tidymodels workflow for predicting flipper length from body mass and bill measurements. Use 5-fold cross-validation to evaluate performance.

7. Build a recipe that includes polynomial features (e.g., squared terms) for bill length and bill depth. Compare this model's performance to a linear model using cross-validation.

8. Use tidymodels to compare at least three different regression models (e.g., linear regression, random forest, support vector machine) for predicting body mass. Which performs best?

9. Create a classification workflow using tidymodels to predict all three penguin species. Calculate and compare precision, recall, and F1 scores for each species.

10. Implement a hyperparameter tuning workflow using `tune_grid()` to optimize the number of trees and minimum node size for a random forest model predicting body mass.

### Advanced Exercises

11. Split the penguin data by island, use one island as a test set, and evaluate how well models trained on other islands generalize. What does this tell you about geographical variation?

12. Create a nested cross-validation workflow: use an outer loop for model evaluation and an inner loop for hyperparameter tuning. Compare this to simple cross-validation.

13. Develop a complete modeling pipeline that includes:
    - Data exploration and visualization
    - Train/test split with stratification
    - Recipe with appropriate preprocessing
    - Multiple model specifications
    - Cross-validation comparison
    - Final model evaluation on test set
    - Interpretation of results

14. Use tidymodels to implement a regularized regression (ridge or lasso) and compare it to ordinary least squares. How does regularization affect the coefficients?

15. Create an ensemble model that combines predictions from linear regression, random forest, and gradient boosting models. Does the ensemble outperform individual models?