---
prefer-html: true
reference-location: document
citation-location: document
execute:
  eval: false
knitr:
  opts_chunk:
    dev: "png"
    python.reticulate: false
---

# R in the Wider Ecosystem

## Introduction

::: {.callout-note}
## Learning Objectives

By the end of this chapter, you will be able to:

1. Choose between RStudio and its alternatives for a given workflow.
2. Move analyses between point-and-click GUIs (jamovi, JASP) and scripted R.
3. Run R from VS Code, Positron, Jupyter, and the terminal.
4. Connect R to Python, databases, and other applications via reticulate, DBI, and plumber.
5. Decide which integrations are mature enough for coursework and which are still experimental.
:::

::: {.callout-warning}
## Version-sensitive chapter

This chapter describes tools whose interfaces, release status, and licensing can change quickly. The examples are intentionally illustrative and are not executed during the normal book build. Check the official documentation and the “last verified” date before using installation commands in a course or production project.
:::

RStudio is not the only way to write R, and R does not work alone. By 2026 the R ecosystem includes a point-and-click GUI used in undergraduate teaching (jamovi), a Bayesian-friendly cousin (JASP), a new flagship IDE from Posit (Positron), polyglot tools that treat R and Python as equals (Quarto, reticulate, VS Code), and a small zoo of ways to expose R to other languages (plumber, Rserve, rpy2). This chapter walks through the integrations that natural-science researchers actually reach for, organised around the question students keep asking: *what should I use, and when?*

::: {.callout-tip}
## PROFESSIONAL TIP: Pick the tool that matches the question

The right environment depends on what you're trying to do, not on what is fashionable:

- Teaching a stats lab to first-year ecology students who fear the command line? Reach for **jamovi**.
- Writing your thesis with mixed R and Python chunks? **Quarto + Positron** (or VS Code).
- Sharing a species-distribution model with a non-R collaborator? Wrap it in **Shiny** or **plumber**.
- Working on a remote HPC cluster over SSH? **Radian** beats the bare R console.

Don't switch tools mid-project unless there's a real payoff. Each switch costs you a day of friction.
:::

## When the GUI is the right answer: jamovi and JASP

### jamovi

[jamovi](https://www.jamovi.org/) is an open-source statistical platform with a spreadsheet-style interface, built on top of R. Students who learned stats in SPSS feel at home immediately. The trick is that every analysis jamovi runs is also an R call to the `jmv` package — you can flip on **Syntax mode** (top-right ⋮ menu) and see the underlying code.

**When to use it.** Introductory teaching, stakeholder demos, or as a sanity check against a scripted analysis.

**Install.**

```bash
# Linux (Flatpak — works on Fedora, Ubuntu, others)
flatpak install flathub org.jamovi.jamovi

# macOS
brew install --cask jamovi
```

**Integration with R.**

```{r}
#| eval: false
# In an R script, reproduce a jamovi ANOVA
library(jmv)
library(jmvReadWrite)

# Read a .omv file directly, with embedded analyses
dat <- jmvReadWrite::read_omv("study1.omv", getSyn = TRUE)

# Or run the same analysis from scratch
jmv::ANOVA(
  data = iris,
  dep  = "Sepal.Length",
  factors = "Species",
  postHoc = ~ Species,
  effectSize = "eta"
)
```

::: {.callout-warning}
`jmv` returns S4 objects, not tibbles. `broom::tidy()` and the rest of the tidyverse will not work on them directly. Treat jmv output as a printed report; if you need a tidy data frame for downstream code, refit the model with `aov()` or `lm()` and tidy that.
:::

### JASP

[JASP](https://jasp-stats.org/) is jamovi's Bayesian-friendly cousin: same point-and-click model, but with strong support for Bayes factors, posterior plots, and meta-analysis. Click the **R Syntax** button on any analysis to see the call.

**When to use it.** First encounter with Bayesian methods, or any analysis where you want a Bayes factor alongside a p-value without writing Stan code.

**Install.**

```bash
# Linux
flatpak install flathub org.jaspstats.JASP

# macOS
brew install --cask jasp
```

::: {.callout-note}
JASP's R syntax export is less complete than jamovi's — some Bayesian modules still don't emit runnable R. Treat the export as a starting point, not a guarantee.
:::

## Editors and IDEs beyond RStudio

### Positron — Posit's new IDE

[Positron](https://positron.posit.co/) is Posit's next-generation IDE, built on Code OSS (the same base as VS Code). Posit has signalled it as the long-term successor to RStudio, while continuing to maintain RStudio in parallel. Releases come weekly; the May 2026 build (`2026.05.x`) is the first I'd recommend for daily writing.

**Why a student would switch.** First-class Python *and* R support side by side. The Variables pane shows both R and Python objects when you use `reticulate`. Quarto preview, Shiny apps, and notebooks all run in-IDE.

**Install.**

```bash
# Linux (Fedora): download the .rpm from positron.posit.co/download.html
sudo dnf install ./Positron-*.rpm

# macOS
brew install --cask positron
```

**What to expect.** RStudio keyboard shortcuts mostly don't carry over — Positron uses VS Code bindings. RStudio addins that depend on the `rstudioapi` package will not run. The trade-off is a much more capable editor with proper multi-language support.

### VS Code with the R extension

If you already write Python or web code in VS Code, you can add R without learning a second IDE.

**Install.**

```bash
# macOS
brew install --cask visual-studio-code

# Then install the R extension and language server
code --install-extension reditorsupport.r
```

```{r}
#| eval: false
install.packages(c("languageserver", "httpgd"))
```

The `httpgd` package gives a live plot pane that updates as you re-run code, which the default VS Code R extension lacks. Add to your `.Rprofile`:

```{r}
#| eval: false
if (interactive() && Sys.getenv("RSTUDIO") == "" &&
    requireNamespace("httpgd", quietly = TRUE)) {
  options(vsc.plot = FALSE)
  httpgd::hgd()
}
```

### Radian — a better R console

[Radian](https://github.com/randy3k/radian) is a Python-based replacement for the bare `R` REPL, with multi-line editing, syntax highlighting, and IPython-style features. Useful when you're SSHed into a server and don't have a graphical IDE.

```bash
pipx install radian
radian
```

::: {.callout-warning}
Radian is no longer under active development as of 2026. It still works, but treat it as legacy rather than a long-term investment. Watch the project's README for a successor recommendation.
:::

## R from a notebook: Jupyter and IRkernel

University HPC platforms and cloud notebooks (Colab-style services, JupyterHub) speak Jupyter, not RStudio. Connecting R is a two-step install:

```{r}
#| eval: false
install.packages("IRkernel")
IRkernel::installspec()  # registers the R kernel with Jupyter
```

```bash
# If Jupyter is not yet on PATH
pipx install jupyterlab
jupyter lab
```

::: {.callout-warning}
Jupyter notebooks (`.ipynb`) diff poorly under git because the output JSON changes every run. Use `jupytext` to pair every notebook with a clean `.R` or `.qmd` file, and put the `.qmd` under version control instead. Reserve notebooks for the final presentation layer.
:::

## R alongside Python: reticulate

The [`reticulate`](https://rstudio.github.io/reticulate/) package lets R call Python — useful when a Python-only library (a remote-sensing wrapper, `scikit-learn`, a satellite-imagery API) is the right tool for one step of an otherwise R pipeline.

```{r}
#| eval: false
library(reticulate)
use_python(Sys.which("python3"))

# Pass an R data frame into Python, train a model, pull predictions back
sk <- import("sklearn.ensemble")
rf <- sk$RandomForestRegressor(n_estimators = 100L)
rf$fit(r_to_py(iris[, 1:4]), iris$Sepal.Length)
predictions <- rf$predict(r_to_py(iris[, 1:4]))
head(predictions)
```

In a Quarto document, R and Python chunks can share variables automatically:

````
```{r}
df <- mtcars
```

```{python}
import pandas as pd
r.df.head()
```
````

::: {.callout-warning}
The single biggest reticulate pitfall is environment management. Pick one approach and stick to it: either `reticulate::virtualenv_create()` and `use_virtualenv()`, or pair `renv` (for R) with `uv` (for Python). Don't mix system Python, conda, and venvs in the same project.
:::

## R talking to other applications

### Databases — DBI and DuckDB

Most labs that maintain a specimen database, environmental monitoring archive, or remote-sensing catalogue store the data in a database. Query it from R rather than exporting CSVs.

```{r}
#| eval: false
library(DBI)
library(duckdb)
library(dplyr)

# DuckDB: a fast analytics database that lives in a single file
con <- dbConnect(duckdb::duckdb(), "specimens.duckdb")
DBI::dbWriteTable(con, "iris", iris)

# dbplyr translates dplyr verbs to SQL transparently
tbl(con, "iris") |>
  group_by(Species) |>
  summarise(n = n(), mean_sl = mean(Sepal.Length)) |>
  collect()

dbDisconnect(con, shutdown = TRUE)
```

For PostgreSQL or MariaDB, swap `duckdb::duckdb()` for `RPostgres::Postgres()` or `RMariaDB::MariaDB()`. The `odbc` package covers commercial databases (SQL Server, Oracle) but requires system-level ODBC drivers — on Fedora, `sudo dnf install unixODBC-devel` first.

### Exposing R as a web service: plumber

When a Python or JavaScript application needs to call an R-only model (a `vegan` ordination, a `brms` Bayesian fit), wrap the R function as a REST endpoint with [plumber](https://www.rplumber.io/).

```{r}
#| eval: false
# In a file plumber.R:
#* @get /predict
#* @param x A numeric value
function(x) {
  list(prediction = as.numeric(x) * 2 + rnorm(1))
}

# Then in R:
plumber::pr("plumber.R") |> plumber::pr_run(port = 8000)
```

Any other language can now call `GET http://localhost:8000/predict?x=3` and get a JSON response. Plumber generates Swagger documentation automatically.

### Shiny — R as an interactive application

Shiny apps embed an R analysis in a browser interface. Useful for sharing a model with a non-R collaborator: the user moves sliders, the app re-runs the analysis on the server, and updated plots appear.

**Shinylive** is the 2024-2026 development worth knowing. It compiles a Shiny app to WebAssembly so the app runs entirely client-side, with no server. Pair it with the `quarto-shinylive` Quarto extension and you can embed a working Shiny app inside a chapter of a book like this one, hosted on plain GitHub Pages.

## Cloud and reproducibility

### Posit Cloud

[Posit Cloud](https://posit.cloud/) (formerly RStudio Cloud) runs RStudio in a browser with zero install. The free tier is still available and useful for students without a working local R install, but as of 2025 new accounts can no longer publish outputs from there — that moved to Posit Connect Cloud. Treat it as a teaching environment, not a publishing platform.

### GitHub Codespaces with Rocker

The [`rocker-org/devcontainer-try-r`](https://github.com/rocker-org/devcontainer-try-r) repository gives a one-click cloud R + RStudio Server environment in GitHub Codespaces. Combined with `r2u` for fast binary package installs, it is the cleanest way I know to ensure 30 students all get the same R setup for an assignment.

### rig — managing multiple R versions

[rig](https://github.com/r-lib/rig) is Posit's R installation manager. It replaces ad-hoc scripts for switching between R versions on a single machine. Worth using once you have more than one R project that pins different versions.

```bash
# Install rig (Linux)
curl -Ls https://github.com/r-lib/rig/releases/download/latest/rig-linux-latest.tar.gz | sudo tar xz -C /usr/local

# Install and use a specific R version
rig add 4.4.3
rig default 4.4.3
```

## A decision guide

The chapter covers many tools. Here is the cheat sheet I would give a student.

| What you want to do | Reach for |
|---|---|
| Teach intro stats with a GUI | jamovi |
| First encounter with Bayesian methods | JASP |
| Modern IDE, R + Python first-class | Positron |
| Already use VS Code, add R | VS Code + R extension + httpgd |
| Cleaner R console in a terminal | Radian (with the maintenance caveat) |
| Run R inside JupyterHub | IRkernel |
| Call Python from R | reticulate |
| Query a database from R | DBI + DuckDB / RPostgres |
| Share an R analysis as a web app | Shiny (or Shinylive in a Quarto book) |
| Expose R to a non-R application | plumber |
| Zero-install R for a classroom | Posit Cloud, or GitHub Codespaces + Rocker |
| Manage multiple R versions | rig |

## Summary

R does not live alone. Each integration in this chapter exists because someone needed R to talk to something else — a GUI for non-coders, a notebook server, a Python library, a web application, a database. None of these tools replaces RStudio for general use; they let you reach contexts where RStudio is not the right answer.

If you take one thing from this chapter, let it be this: a good analytical workflow is whichever combination of tools lets you do your science reproducibly, share your results, and not lose a day of work to environment problems. The R ecosystem is plural by design — use that.

## Exercises

1. Install jamovi. Load a built-in dataset, run a one-way ANOVA, switch to Syntax mode, and copy the generated `jmv::ANOVA()` call into an R script. Confirm you get identical output.
2. Install Positron. Open this book's repository in it and render a single chapter from the Quarto pane. Note three differences from RStudio that affect your workflow.
3. In an R session, install `reticulate` and create a virtual environment. Run a small `scikit-learn` regression on `mtcars`, return predictions to R, and plot them with `ggplot2`.
4. Write a plumber endpoint that takes a species name (string) and returns the mean body size from the Palmer penguins dataset for that species as JSON. Test it with `curl`.
5. Set up `IRkernel` and open a Jupyter notebook. Reproduce one analysis from Chapter 5 in the notebook. Then convert the notebook to a `.qmd` file with `quarto convert`.
6. Connect to a DuckDB database holding a CSV from `data/`. Use `dbplyr` to filter and summarise without `collect()`ing first. Inspect the generated SQL with `show_query()`.

## References

- Positron documentation: <https://positron.posit.co/>
- jamovi documentation: <https://docs.jamovi.org/>
- JASP documentation: <https://jasp-stats.org/>
- reticulate site: <https://rstudio.github.io/reticulate/>
- plumber site: <https://www.rplumber.io/>
- IRkernel installation: <https://irkernel.github.io/installation/>
- Rocker devcontainers: <https://github.com/rocker-org/devcontainer-try-r>
